Progressive postnatal hearing development limits early parent-offspring vocal communication in the zebra finch

Version 1

Abstract

Acoustic communication relies critically on the receiver’s ability to hear. In precocial bird species hearing can already be functional during embryonic stages in the egg, while in altricial bird species hearing typically starts after hatching. Recent research suggests that zebra finch embryos, despite being altricial, have functional hearing already in the egg to engage in parent-embryo acoustic communication and in anthropogenic noise detection. However, their auditory sensitivity during early development remains unknown. Here, we measure auditory brainstem responses over early postnatal development and show that zebra finch hatchlings are deaf even to loud, broadband sounds with hearing responses emerging between 4-8 days after hatching. Auditory sensitivity develops progressively and reaches adult levels by day 20, contradicting the notion of early parent-to-embryo communication in zebra finches. Additionally, egg vibrations induced by sound remain far below detection thresholds of vibrotactile senses. The striking timing coincidence between maturation of the peripheral auditory system and the onset of song learning suggests that hearing functionality may gate the onset of vocal learning in zebra finches.

Introduction

The ability to detect sound is critical for vocal communication, species recognition, predator detection, and survival across the vertebrates (Nordeen and Nordeen, 1992; Heaton et al., 1999; Zevin et al., 2004; Ladich and Winkler, 2017). In birds, precocial species that are able to move around after hatching show some hearing function already inside the egg (Gottlieb, 1965; Saunders et al., 1973; Konishi, 1973). This enables embryo-to-embryo (Vince, 1966; Woolf et al., 1976; Schwagmeyer et al., 1991; Noguera and Velando, 2019) and parent-to-embryo communication that is critical to imprinting on species-specific sounds (Gottlieb, 1965; Grier et al., 1967; Hess, 1972). In contrast, hearing function in altricial birds is thought to mature predominantly after hatching with little to no functional hearing during embryonic stages (Woolley, 2017).

Remarkably, recent research directly challenges this concept and suggests that altricial song-birds can detect soft sounds while still in the egg (from here onwards referred to as embryos). Zebra finch embryos supposedly are epigenetically guided to adapt to high temperatures by their parents high frequency “heat calls” (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023) (from here onwards referred to as heat whistles) and can detect anthropogenic noise as prehatching noise exposures reduce embryonic survival and reproductive success in adulthood (Meillère et al., 2024). Furthermore, fairy wren embryos supposedly learn elements of the female incubation call already at 10-12 days after fertilization (Colombelli-Négrel et al., 2012, 2014, 2016; Kleindorfer et al., 2024). The above interpretations critically depend on the ability of these embryos to detect soft level sound in the egg and after hatching. However, it is currently unknown whether these species can detect sound this early in their development.

The zebra finch is a major animal model system for studying the neural circuitry underlying hearing-guided behaviours including vocal imitation learning. However, we know surprisingly little about development and function of the auditory organ in this species. In singing males, the onset of sensory and sensorimotor learning starts around 25 days-post-hatch (DPH) and continues to adulthood at ~90 DPH (Immelmann, 1967; Böhner, 1990; Roper and Zann, 2006; Braaten, 2010; Gobes et al., 2019). Aligned with the onset of this behaviour, the overall hearing sensitivity is comparable to adult levels at 20 DPH with a frequency range from 0.25 to 8 kHz and highest sensitivity around 3 kHz (Amin et al., 2007; Okanoya and Dooling, 1987; Yeh et al., 2023). However at 10 DPH – the earliest age measured – hearing sensitivity is reduced by 25 dB compared to 20 DPH and is thus ~20-fold lower (Amin et al., 2007). Because hearing sensitivity in zebra finches increases progressively in late postnatal development - as in other bird species (Rebillard and Rubel, 1981; Katayama, 1985; Jones et al., 2006; Brittan-Powell and Dooling, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017) - their ability to detect sounds as embryos and as early hatchlings may be severely compromised. Furthermore, recent work shows that heat whistles are extremely soft sounds in vivo (~34 dB re. 20 µPa at 10 cm) (Anttonen et al., 2025). Therefore, zebra finch embryos would require highly sensitive hearing function to detect such soft sounds. However, we lack conclusive evidence on the hearing capabilities of juvenile zebra finches.

Here we test the hypothesis that zebra fiches can hear during early postnatal development by measuring auditory brainstem responses (ABRs) that allow hearing threshold estimation when behavioural training is not possible (He et al., 2008). Furthermore, we test whether loud sound fields can induce egg vibrations of sufficient magnitude for vibrotactile detection. We show that the earliest ABRs evoked by intense, broadband sounds could not be detected on 2 DPH, but earliest at 4-8 DPH. Responses to frequency specific sounds develop later in a progressive, low-to-high frequency manner. ABR sensitivity to clicks matures to adult-like levels at 10 DPH but the response amplitude continues to increase beyond 20 DPH. Airborne sounds induce low amplitude vibrations in eggs that are far below typical avian and mammalian vibrotactile detection thresholds. Together, our findings show that zebra finch embryos and early hatchlings are functionally deaf and provide evidence against the hypothesis that they can detect sound stimuli via hearing or via vibrotactile senses. Our data furthermore show that auditory nerve responses continue to mature past 20 DPH, suggesting that the maturation of hearing limits the onset of vocal learning in zebra finches.

Results

Responses to loud clicks appear during the first postnatal week

To assess the development of zebra finch hearing sensitivity in early postnatal development we quantified the responses of auditory pathway neurons to sound stimuli by recording ABRs (Fig. 1A). First, to evaluate overall hearing sensitivity we presented loud yet physiologically relevant (95 dB SPL) clicks that stimulate the auditory system across a wide frequency spectrum (Fig. 1B). Clicks activate most auditory neurons nearly simultaneously and result in prominent ABRs, thus providing maximal sensitivity for threshold detection.

Click-evoked ABRs appear and gradually mature after hatching in zebra finches.

(A) ABR recording setup. (B) Used sound stimuli. Red, phase inverted signal. (C) Representative, 95 dB SPL click-evoked ABR ante- (black) and post-mortem (red). (D) Representative ABRs to 95 dB SPL clicks at tested DPH timepoints. The first negative peak (I) and the following trough are designated as wave I. Noise RMS is determined from the red area and signal RMS from the blue area. (E) Percentage of animals that exhibited a detectable (S/N > 2) click-evoked ABR. (F) Detection thresholds were measured with a set of clicks with decreasing 5 dB SPL steps. Arrow indicates the determined threshold for this 8 DPH example. (G) Average click thresholds at different DPH ages. Orange box indicates heat whistle playback levels used in previous studies (Katsis et al., 2018). Red box indicates in vivo heat whistle levels at 10 cm distance (Anttonen et al., 2025). (H) Click-evoked ABR wave I amplitude and (I) latency over SPL. Age groups are color-coded in D-I. Shaded areas in H-I = s.e.m.

Clicks induced robust neural potentials in adult zebra finches (Fig. 1C-E). These potentials were not recording artefacts as they occurred only when the animal was alive (N = 3, Fig. 1C). None of the tested animals displayed click-evoked ABRs at 2 DPH (both with detection at signal-to-noise ratio of 2 (S/N > 2) and with detailed visual inspection; n=0/15; Fig. 1D,E). However, click-evoked ABRs became observable in some animals during the first postnatal week (S/N > 2; 4 DPH, n = 5/13; 6 DPH, n = 9/10; Fig. 1D,E). Subsequently, all tested animals consistently demonstrated click-evoked ABRs from the second postnatal week onwards (S/N > 2; 8 DPH, n = 11; 10 DPH, n = 11; 20 DPH, n = 7; 25 DPH, n = 8; Adult, n = 8; Fig. 1D,E). Thus, sounds at loud, yet physiologically relevant SPLs do not evoke ABRs in the first days after hatching and in the majority only after one week.

Click-evoked ABR thresholds are adult-like at 10 DPH

To estimate hearing sensitivity thresholds over postnatal age, we recorded series of click-evoked ABRs at 5 dB downward steps from 95 dB SPL (Fig. 1F). Hearing thresholds are strongly affected by the age of the animal (F(6) = 25.22, p = 4.93*10−14; Fig. 1G, Table S1). Click-evoked ABR thresholds are significantly higher than adult-like levels (40.6 ± 4.0 dB SPL) in animals younger than 10 DPH but not in older animals (4 DPH vs. adult, t = 8.16, p = 3.27*10−10; 6 DPH vs. adult, t = 5.92, p = 1.39*10−6; DPH 8 vs. adult, t = 3.37, p = 0.008; 10 DPH vs. adult, t = 1.23, p = 1; 20 DPH vs. adult, t = −0.13, p = 1; 25 DPH vs. adult, t = −1.29, p = 1; p-values are Bonferroni-corrected; Fig. 1G, Table S1). Click thresholds are as high as 80.6 ± 1.6 dB SPL on 4 DPH. Thus, zebra finch hearing sensitivity increases rapidly but gradually during the first two postnatal weeks.

ABR wave I gradually reaches maturity 25 days after hatching

The click-evoked ABR consists of multiple wave peaks and troughs that reflect the subsequent activations of various neural populations of the auditory pathway. The first wave (wave I in Fig. 1D) is considered to correspond to the activity of the primary auditory neurons (Buchwald and Huang, 1975). To study the postnatal maturation of this synchronized neural output, we measured changes in both the amplitude and the latency of wave I over postnatal development.

In adults, the amplitude of wave I increases significantly with click SPL (Fig. 1H). Age significantly affects both wave I amplitude directly and the relation between SPL and wave I amplitude (Age-SPL interaction compared to only additive effects of age and SPL: χ2(6)=164.23, p=7.52*10−33, additive effects of age and SPL compared to only SPL: χ2(6)=79.43, p=4.69*10−15, SPL only compared to null model: χ2(1)=438.12, p=2.78*10−97, Fig. 1H, Table S2). The steepness of the amplitude-SPL slope gradually increases with age and is significantly different between adults and younger animals, but not between adult and those aged 25 DPH (4 DPH vs. adult, t = −5.40, p = 1.50*10−6; 6 DPH vs. adult, t = −5.65, p = 1.8910−6; DPH 8 vs. adult, t = −6.30, p = 3.79*10−7; 10 DPH vs. adult, t = −5.53, p = 6.53*10−6; 20 DPH vs. adult, t = −3.00, p = 0.03; 25 DPH vs. adult, t = −0.69, p = 1; p-values are Bonferroni-corrected; Fig. 1H, Table S2).

The latency of wave I decreases with increasing click SPL (Fig. 1I). Furthermore, age significantly affects the relation between dB SPL and wave I latency (Age-SPL interaction compared to only additive effects of age and SPL: χ2(6) = 116, p = 1.31*10−22; additive effects of age and SPL compared to only SPL, χ2(6) = 109, p = 2.76*10−21; SPL only compared to null model: χ2(1) = 754, p = 4.81*10−166; Fig. 1I, Table S3). The observed latencies are significantly longer during early postnatal development (4-to-10 DPH) when compared to latencies observed in adults. No significant differences are observed between the latencies of adults and animals older than 10 DPH (4 DPH vs. adult, t = 12.56, p = 3.41*10−19; 6 DPH vs. adult, t = 10.06, p = 1.49*10−13; 8 DPH vs. adult, t = 7.24, p = 1.16*10−8; 10 DPH vs. adult, t = 6.18, p = 5.95*10−7; 20 DPH vs. adult, t = 2.24, p = 0.18; 25 DPH vs. adult, t = 1.67, p = 0.61; p-values are Bonferroni-corrected; Fig. 1I, Table S3).

Taken together, the synchronized output of the auditory nerve increases in sensitivity and reduces in latency postnatally progressively from 4 DPH until reaching maturity at 20-25 DPH. Such gradual increase is consistent with postnatal hearing development in budgerigars, mice, and humans (Brittan-Powell and Dooling, 2004; Hecox and Galambos, 1974; Henry and Haythorn, 1978).

Hearing sensitivity develops progressively from low to high frequencies

When in a hot environment, zebra finch parents emit heat whistles that have been proposed to epigenetically guide their unhatched offspring to adapt to high temperatures (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023). These whistles have been reported to have high fundamental frequencies (7-10 kHz) when compared to zebra finch vocalisations (Mariette and Buchanan, 2016). As hearing sensitivity typically develops in a low-to-high frequency order (Brittan-Powell and Dooling, 2004; Kraemer et al., 2017; Ehret, 1976; Ehret and Romand, 1981), such high frequency sounds may be challenging to be detected by the immature auditory system.

To study the postnatal development of frequency-specific hearing sensitivity we tested whether loud (95 dB SPL) tone bursts centred at frequencies between 0.25-to-8 kHz elicit detectable ABRs. Tone burst-evoked ABRs are first detectable on 8 DPH at 1 kHz (Fig. 2A and Fig. S1A). As observed with click stimuli, the amplitude of tone burst-induced ABRs increases with age (Fig. 2A, Fig. S1B, Table S2; Age as a predictor of amplitude compared to null model: χ2(3) = 6.11, p = 0.11). The latency of the response decreases significantly with age (Fig. 2a, Fig. S1c, Table S2; Age as predictor of latency compared to null model: χ2(3) = 33.72, p = 2.27*10−7). Latencies are significantly longer at 8-to-10 DPH than at adulthood (1 kHz: 8 DPH, 6.60 ± 0.77 ms; 10 DPH, 5.01 ± 0.40 ms) and are adult-like at 20-to-25 DPH (1 kHz: 20-to-25 DPH, 3.34 ± 0.18 ms; adult, 3.67± 0.30 ms; 8 DPH vs. adult, t = 5.95, p = 1.87*10−6; 10 DPH vs. adult, t = 4.52, p = 0.0004; 20-to-25 DPH vs. adult, t = 1.16, p = 0.81; p-values are Bonferroni-corrected; Fig. S1C and Table S2). During the following weeks of postnatal development, tone burst-induced ABRs became detectable at 0.5-to-4 kHz (Fig. S1A). No responses were detected at higher frequencies (Fig. 2B,C; Fig. S1A). Taken together, the maturation of hearing sensitivity in zebra finches progresses in a low-to-high frequency order as is also observed for example in mice (Ehret, 1976), cats (Ehret and Romand, 1981), and in other birds (Brittan-Powell and Dooling, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017).

Sound frequency sensitivity matures over postnatal development.

(A) Representative ABRs to 1 kHz and (B) 8 kHz tone burst stimuli over postnatal development. (C) Zebra finch ABR thresholds obtained with tone bursts (solid lines) and with tone pips (dashed line). Dotted line displays a behavioural threshold reported by Okanoya and Dooling (1987). Shaded areas = s.e.m. The level of heat whistle playbacks used in previous studies (Katsis et al., 2018) (orange box) and the level of in vivo heat whistles at 10 cm distance (Anttonen et al., 2025) (red lines) are just on or below the adult hearing threshold.

In vivo heat whistles are below adult hearing threshold levels

Most birds have strongly decreasing hearing sensitivity above 5 kHz (Okanoya and Dooling, 1987; Yeh et al., 2023; Rebillard and Rubel, 1981; Brittan-Powell and Dooling, 2004; McGee et al., 2019). Given the 7-10 kHz frequency range of heat whistles, it is unknown if even adult zebra finches can readily detect them.

To test this, we analysed adult hearing sensitivity for higher frequency sound stimuli in more detail. Importantly, tone bursts evoke ABRs only in about half of the adults (Fig. S1A) and the detection thresholds of > 60 dB SPL are high. This indicates that while being more frequency specific, tone bursts failed to recruit auditory neurons to fire concurrently enough to generate detectable ABRs. Therefore, we next used shorter 5 ms-long tone pips (Fig. 1B) that have a broader spectral distribution of acoustic energy compared to tone bursts and produce a more focused temporal response (Lauridsen et al., 2021). Indeed, the SPL matched tone pips produced much more prominent ABRs (Fig. 2C and Fig. S2A) that were readily detected in all tested adults around their most sensitive hearing range of 1-to-4 kHz (Amin et al., 2007; Okanoya and Dooling, 1987; Yeh et al., 2023) (Fig. S2B). Detection thresholds for tone pips are significantly more sensitive than those of tone bursts with ~20 dB SPL (Additive effects of stimulus type and frequency as predictors of threshold compared to stimulus type alone χ2(1) = 6.84, p = 0.009; stimulus type as a predictor of threshold compared to null model: χ2(1) = 10.27, p = 0.001, Table S2). Importantly, the tone pip-based audiogram has similar shape and is shifted ~10-to-20 dB SPL upwards compared to behavioural thresholds (Okanoya and Dooling, 1987) (Fig. 2C), which is consistent with previous studies (Stapells, 2000).

Based on these audiograms, we can conclude that even for adult zebra finches, heat whistles at playback levels of 62-to-67 dB SPL at the ear (Katsis et al., 2018) and at in vivo levels of 33.9±3.3 dB SPL at 10 cm (Anttonen et al., 2025) are just on or below their detection limit. More importantly, as hearing sensitivity is much lower in less developed hatchlings (Fig. 1 and 2), these data further suggest that early hatchlings cannot hear high-frequency heat whistles.

Sound-induced egg vibrations are below detectable levels

Because our data strongly suggests that embryonic zebra finches cannot detect soft, high frequency heat whistles as airborne sounds via their auditory system, we next tested if airborne sounds can induce zebra finch eggs to move at sufficient velocities to be detected as vibrations (Håkansson et al., 1985; Sakai et al., 2006; Bergin et al., 2015; Chhan et al., 2017; Mountcastle et al., 1972; Hörster, 1990). We played sound frequency sweeps to eggs while simultaneously recording the sound-induced vibrations of the egg with a laser vibrometer (Fig. 3A). The vibration velocity transfer function shows that eggs vibrate at velocities around −85 to −55 dB re 1 mm s−1 when exposed to a loud 94 dB SPL frequency sweep between 0.25-to-10 kHz (Fig. 3B). Within the frequency ranges associated with the most sensitive hearing of zebra finches (1-4 kHz, green) and with heat whistles (7-10 kHz, red), sound exposures at 94 dB SPL induce average vibration velocities of −72.4 ± 0.8 and −69.2 ± 0.9 dB re 1 mm s−1, respectively (Fig. 3B,C). We calculated the corresponding vibration velocity levels induced by SPLs used during experimental heat whistle playbacks to zebra finch eggs (62-to-67 dB SPL) (17) to be between −99.4 and −96.2 dB re 1 mm s−1, respectively (Fig. 3D). The obtained vibration velocities are approximately 10-to-40 dB lower than typical vertebrate bone conduction thresholds (Håkansson et al., 1985; Sakai et al., 2006; Bergin et al., 2015; Chhan et al., 2017) and 90 dB lower than what human (Mountcastle et al., 1972) and bird (Hörster, 1990) vibrotactile senses typically can detect (Fig. 3D). Detection of in vivo heat whistles is even less likely as they are at least 27 dB softer compared to the playbacks levels (Anttonen et al., 2025). Inconclusion, air-borne sounds at physiologically relevant SPLs are not able to induce zebra finch eggs to vibrate at high enough velocities to be detected by known avian and mammalian vibrotactile senses.

Intense sound stimulation induces zebra finch eggs to vibrate at very low velocities.

A Setup used to measure sound induced vibrations of eggs. B Average (N=5) vibration velocity transfer function of a zebra finch egg during a 94 dB SPL sound sweep between 0.25-to-10 kHz. Gray area marks the standard deviation. Green and red areas display the frequency windows related to the best hearing sensitivity of zebra finches and to heat whistles, respectively. C The average egg vibration velocities induced by a 94 dB SPL sounds at frequencies associated with song (green) and with heat whistles (red). Error bars = s.d. D The average egg vibration velocities caused by 67 dB SPL sounds plotted together with thresholds for human (Håkansson et al., 1985) and mouse bone conduction hearing (Chhan et al., 2017) and for human vibrotactile threshold at 0.2 kHz (Mountcastle et al., 1972).

Discussion

We show that zebra finch early hatchlings are functionally deaf, and that hearing functionality gradually increases over postnatal development. The synchronized output of the auditory nerve is first detectable at 4-8 DPH after which hearing sensitivity gradually increases until reaching adult levels at 20-25 DPH. This increase in hearing sensitivity over postnatal development from low-to-high frequencies is found in both altricial and precocial bird species (Saunders et al., 1973; Konishi, 1973; Rebillard and Rubel, 1981; Katayama, 1985; Jones et al., 2006; Brittan-Powell and Dooling, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017; Aleksandrov and Dmitrieva, 1992; Lippe, 1994) including zebra finches (Amin et al., 2007) - as well as in mammals (Henry and Haythorn, 1978; Ehret and Romand, 1981,?). Therefore, the functional maturation of hearing in zebra finches follows a gradual, progressive pattern consistent with other birds and mammals.

The progressively increasing hearing sensitivity over postnatal development (Fig. 1) strongly suggests that zebra finch embryos have an even lower hearing sensitivity than early hatchlings. Even in precocial bird species that show auditory function already in the egg, auditory thresholds remain very high until a few days before hatching (Saunders et al., 1973; Rebillard and Rubel, 1981; Lippe, 1994; Saunders et al., 1974). In precocial chickens, evoked responses to very loud sounds (>100 dB SPL) have been obtained around embryonic day 12 (chickens hatch at day 21) (Saunders et al., 1973), but ambient sound levels evoke responses around four days later (Jones et al., 2006). Importantly, these measurements involved experimentally opening the egg, clearing the auditory meatus from fluids, and even drainage of fluids from the middle ear (Rebillard and Rubel, 1981) and thus may overestimate the natural hearing sensitivity. Although ABR provides easily accessible measurements of hearing sensitivity, the technique does not inform on the anatomical and neural processes underlying the gradual development of hearing sensitivity. In birds, these processes have been studied in most detail in chickens, where fluid unloading and maturation of middle ear structures causes significant improvements in hearing sensitivity during development (Saunders et al., 1973; Cohen et al., 1992). The onset of evoked responses to very loud sounds furthermore coincides with synapse formation on hair cells of the inner ear (Saunders et al., 1973; Cohen and Fermin, 1978; Caus Capdevila et al., 2021). Furthermore, the response characteristics of the auditory nerve fibers and the latencies of the ABR show postnatal changes (Katayama, 1985; Jones et al., 2006; Manley et al., 1991). The periphery of the chicken auditory system also continues to anatomically mature for several weeks after hatching (Cohen et al., 1992; Ryals et al., 1984; Cotanche and Corwin, 1991). When these hallmark events occur in zebra finch development remains unknown. Further work is required to understand the physiological basis for postnatal development of hearing sensitivity in the zebra finch.

Our data provides evidence against the hypothesis that zebra finch embryos and early hatchlings can detect vibroacoustic stimuli via the auditory system or via vibrotactile senses. Our results therefore challenge previous interpretations that zebra finch parents use sound or vibration to communicate with their hatchlings and embryos (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023). Particularly, the late maturation of high frequency hearing makes heat whistles (7-10 kHz) challenging to be detected by adults, let alone by developing zebra finches. The difficulty to detect high frequency signals is not unique to zebra finches; even precocial species that have hearing function prior to hatching have high response thresholds for sounds above 6 kHz as embryos (chicken: >120 dB SPL; mallard duck: >90 dB SPL) (Saunders et al., 1973; Katayama, 1985). The established parent-embryo communication in precocial mallard ducks depends specifically on the low-frequency auditory sensitivity of the embryo to detect maternal calls with frequency components below 3 kHz (Gottlieb, 1975) which is consistent with limited high frequency hearing. An alternative explanation for the detection of heat calls by zebra finch embryos (Mariette, 2024) could be sound-induced vibrations of the egg or bone conduction hearing. However, even if sound-induced vibrations of the egg would translate without loss into vibrations of the embryo, our data show that the generated vibrations are orders of magnitudes too small to evoke responses from typical vertebrate auditory or vibratory senses (Mountcastle et al., 1972; Hörster, 1990; Jones and Jones, 1996; Freeman et al., 1999). Taken together, our results provide evidence against the hypothesis that zebra finch embryos can hear sounds, and we consider it thus unlikely that adult zebra finches can communicate with their embryos via air-borne sound. How the heat whistle and anthropogenic noise playbacks have affected the embryos in the earlier studies (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023; Meillère et al., 2024) remains to be further tested. Importantly, the earlier work with heat whistle playbacks must be reproduced at physiologically relevant sound levels as they are ~27 dB too high when compared to in vivo heat whistles (Anttonen et al., 2025).

The postnatal maturation of hearing may also affect the onset and efficacy of vocal imitation learning. In zebra finches, sensory song learning starts on 25 DPH (Immelmann, 1967; Böhner, 1990; Roper and Zann, 2006; Braaten, 2010; Gobes et al., 2019). While the auditory sensitivity to click stimuli appears be mature at 10 DPH, the response amplitude of the auditory nerve (ABR wave I) increases past 20 DPH. Furthermore, ABR features that are generated by more upstream neural populations, e.g. wave II and later, appear to go through major changes after 10 DPH (Fig. 1C). These data suggest that both peripheral auditory nerve and downstream circuits of the zebra finch auditory system continue to develop for several weeks after hatching. In chickens, the temporal coding ability of the auditory nuclei the ability to respond to high stimulus rates – also increases post-hatch (Saunders et al., 1973, 1974; Manley et al., 1991; Warchol and Dallos, 1990). However, we know surprisingly little about these changes in zebra finches in early development (Hong and Sanchez, 2018) but they seem to continue well into the sensorimotor phase of song learning. The striking co-occurrence of the onset of the sensory phase of song learning with the maturation of the auditory nerve responses suggests that the onset of song learning in zebra finches may be gated by the developmental maturation of the auditory system. Furthermore, nascent, internal song templates acquired prior to the full maturation of hearing function may also require updating to match the changing target.

Methods and materials

Animals

Local, captivity-bred zebra finches (Taeniopygia guttata, order Passeriformes) were used in this study. Zebra finches were kept and bred indoors in temperature- and humidity-controlled group aviaries (100-to-200 individuals) or pairwise in breeding cages supplemented with nesting boxes. Lights in the aviaries were on a 12/12 hour light/dark photoperiod. Zebra finches were given water, food, cuttlefish bones, and nesting material ad libitum. Nesting boxes were monitored daily for accurate aging of the hatched chicks. The day when a given zebra finch chick was observed to have hatched in the morning was defined to be post-hatch day 0 (0 DPH). Hatched chicks were recognized from each other by making individualized, non-invasive down feather cuts until the usage of leg bands was feasible at DPH 10. Animal experiments were approved by the Danish National Animal Experimentation Board (Dyreforsøgstilsynet) under license number 2019-15-0201-00284.

Anesthesia

Preparation

30-to-60 minutes prior to recording, 20 DPH-to-adult zebra finches were removed from their breeding cage or aviary, brought to the test room in a transport cage, and were fasted to minimize the reflux of fluid/food during anesthesia. Younger zebra finches were moved to the test room in a warmed transport box and were not fasted. Zebra finches were weighed prior to the induction of anesthesia.

Dosing

Anesthesia was induced with an intrapectoral injection of ketamine (Ketaminol Vet., 100 mg/ml, MSD Animal Health, NJ, USA) in combination with xylazine (Rompun vet., 20 mg/ml, Bayer Animal Health GmbH, Leverkusen, Germany). 20 DPH-to-adult zebra finches were administered 50 µg of ketamine and 10 µg of xylazine per gram of body weight. 2-to-10 DPH zebra finches were given 50 µg of ketamine and 5 µg of xylazine per gram of body weight. After one hour, a supplementary dose (75% of the induction dose) was administered. In some zebra finches, a clear liquid - possibly due to fluid reflux or due to increased production of oral secretions - was noticed in their mouths within 10-to-15 minutes from anesthesia induction. To avoid airways from being obstructed, a small piece of paper tissue was inserted to the side of the mouth to drain the liquid.

Auditory brainstem responses (ABRs)

Recording room

All recordings were performed by placing the zebra finch inside an audiometric room (T-room, C-A TEGNÉR AB, Bromma, Sweden) equipped with a self-built vibration isolation table with a padded surface. The walls of the room were padded with sound absorbing wedges (Classic Wedge 30, EQ Acoustics, UK).

Maintenance of body temperature

The body temperature of the animals was maintained by placing a heating pad warmed by circulating water (TP702, Gaymar Instruments Inc., Orchard Park, NY, USA) underneath them and placing a small cloth on top of their body. Body temperature monitoring was performed with a J-type temperature probe connected to a thermocouple measurement device (NI USB-TC01, National Instruments). The tip of the probe was coated with Sylgard (184 Silicone Elastomer Kit, Dow Europe GMBH, Wiesbaden, Germany), covered with vaseline, and was placed withing the cloaca. With 2-to-10 DPH zebra finches where the cloaca was too small for probe insertion the probe was placed under the wing.

Head placement

The head of the zebra finch was slightly elevated by placing it on top of wax mound (Surgident Periphery Wax, Kulzer, IN, US). The ends of a wedge-shaped notch on the wax were slightly pressed together so that the wax held the lower mandible of the zebra finch in place. The auricular feathers covering the left external auditory meatus in mature zebra finches were cut with scissors to allow unobstructed stimulation of the auditory system. Prior to electrode placement, the appropriate depth of anaesthesia was confirmed by the lack of responsiveness to a toe-pinch.

Electrodes

Three subdermal needle electrodes were placed subcutaneously on the head of the animal in the following configuration (Fig. 1A): (1) Active electrode: Dorsal to the left auditory meatus, pointing towards to the top of the head. (2) Inverting electrode: On top of the head along the midline, pointing towards the bill; (3) Ground electrode: Dorsal to the right auditory meatus, pointing towards to the top of the head. Of note, to compare these recordings to studies where the positions of the active and inverting electrodes are the opposite one must invert the recorded signal. The impedances of the electrodes were measured (RA4LI, Tucker-Davis Technologies, FL, USA) to be below 1 kOhm in 2-to-10 DPH zebra finches and below 3 kOhm in 20 DPH-to-adult zebra finches.

Recording devices

Recorded evoked potentials were passed from the electrodes to a headstage (RA4LI) in series with a preamplifier (RA4PA) and with a digital signal processor (RM2, all from Tucker-Davis Technologies, FL, USA). RM2 was controlled with custom-written software QuickABR (Brandt et al., 2018). Recorded responses were digitized with a sampling rate of 25 kHz (16 bits).

Sound stimulation

Sound stimulation was generated in the RM2 controlled by QuickABR, amplified (Cambridge Audio, Azur 740A Integrated Amplifier, London, UK), and emitted from a loud-speaker (Wharfedale Diamond 220, Wharfedale Ltd., Huntingdon, UK). The loudspeaker was string-suspended 25 cm away from the animal, directly facing the left auditory meatus. Sound reflections were minimized by placing the loudspeaker at a 45° angle relative to the closest wall of the recording room. To calibrate the loudspeaker, a½” microphone (G.R.A.S. microphone type 26AK, G.R.A.S., Holte, Denmark) was suspended at the position of the head of the animal and connected to a microphone amplifier (G.R.A.S. Power Module Type 12AA, Gain 0, Linear filtering, Direct mode select = Ampl.). The microphone itself was calibrated using an acoustical calibrator (type 4321, Output: 1 kHz at 94 dB re. 20 µPa, Brüel & Kjær). Calibration of the loudspeaker was controlled by QuickABR.

Stimuli

ABRs were elicited with broadband clicks (20 µs) and 25 ms-long tonebursts at 0.25 kHz, 0.5 kHz, 1 kHz, 2 kHz, 3 kHz, 4 kHz, 6 kHz, and 8 kHz. To minimize spectral splatter caused by abrupt on/offsets of stimuli, the rise/fall time (5 ms, cos2-gated) of the tonebursts was selected to accommodate at least one full cycle of the stimulus with the lowest frequency. To further increase the frequency specificity of the stimulus, the plateau time (15 ms) was selected to allow at least 3 full cycles of the stimulus with the lowest frequency. Additionally, 5 ms-long tonebursts (1 ms rise/fall time, cos2-gated) were collected from adult zebra finches. Stimuli were sampled at 25 kHz and repeated at a rate of 25 Hz to collect average responses consisting of 400 individual responses. Every second presented individual stimulus was inverted to reduce microphonic effects. The maximal sound pressure level (SPL) used in this study was selected to be 95 dB SPL as it can be viewed as being physiologically relevant. Click amplitudes are stated as the root mean square (RMS) dB value of a sinusoid with the same peak amplitude as the click (peak equivalent dB).

Recording paradigm

The stability of the recorded responses was monitored throughout the recording session by recording a 95 dB SPL click-evoked ABR periodically between each stimulus type. A specific stimulus recording was accepted into further analysis if the amplitude of the first click-evoked ABR wave had changed less than 25% from the most temporally stable amplitude of that recording session. To minimize test order bias, the recording order of stimuli with varying frequencies was randomized. Three animals were killed with an overdose of ketamine at the end of the recording session to verify the lack of response potentials post mortem.

Egg vibration measurements with laser doppler vibrometry

Zebra finch eggs were removed from their nest and placed in an upright position on a dental wax covered platform at the centre of an anechoic room. The anechoic room was custom-made with its walls and floor covered with 30 cm rockwool wedges. The sample platform was mounted on a heavy tripod which also held a laser Doppler vibrometer (OFV-5000 with OFV-505 sensor, Polytec, Waldbronn, Germany) attached on a manipulator. The laser was pointed at the centre of the side of the egg. Strong reflections were directly obtained from the egg surface so adding reflectors was not necessary. Sound-induced vibrations of the egg were induced with a speaker (JBL IG, Northridge, CA, USA) placed next to the laser vibrometer. A microphone (½ inch G.R.A.S., Brüel & Kjær) was placed few millimeters above the egg and was used for speaker calibration. The egg was stimulated with sound frequency sweeps (0.25-to-10 kHz) at 84 dB SPL. The sweeps were generated with Tucker-Davis system 2 hardware (Tucker-Davis Technologies, TDT, Alachua, FL, USA) controlled by custom-written software DragonQuest (Christensen-Dalsgaard and Manley, 2008). The signal was deconvoluted by dividing the spectrum of the sweep with the measured transfer function of the speaker. The analog laser signal of the laser vibrometer was digitized using an A–D converter (AD2, TDT) and averaged over 100 presentations. During recordings, the laser signal was monitored using an oscilloscope and a vibrometer controller (Polytec OFV-5000). Stimulation and recording were controlled by DragonQuest.

Data analysis and visualization

Data was visualized and analysed with custom scripts in MATLAB R2022B (Mathworks). Figures were edited with Illustrator CS6 (Adobe).

Threshold determination

For ABR threshold detection, the stimulus level was lowered from the maximal value in 5 dB steps. For the reported thresholds to reflect the average sensitivity of zebra finches capable of detecting the given stimulus, only zebra finches that displayed a detectable ABR (both visually and above the signal-to-noise ratio of 2) at the maximal stimulation level are included in the analysis. The response threshold was designated to be 2.5 dB below the last observable response-generating stimulus level (arrow in Fig. 1F).

Signal-to-noise ratio calculation

The recording prior to stimulus onset was used to calculate the noise RMS. Signal-to-noise ratio (S/N) was determined by dividing the RMS value of the signal (blue window) by the RMS of the recording prior to stimulus onset (red window, Figs. 1D; 2A,B and Fig. S2A).

Response latency and amplitude calculation

Response latency was calculated as the time difference between the first negative deflection of the ABR (designated as wave I) and the arrival time of the sound at the ear (Fig. 1c). The amplitude of the wave I (reflecting the activity of the auditory nerve) was calculated as the potential difference between the first negative peak and the following positive trough that together formed the most prominent component of the response (Fig. 1c). Of note, a second negative peak/deflection between the first peak and the trough existed. This peak appeared to merge with the first negative peak at soft SPLs as reported for ABRs of bald eagles and red-tailed hawks (McGee et al., 2019).

Statistics

To test for main effects, we used linear mixed effects models. We sequentially compared models by adding hypothesised effects as fixed effects one-by-one, starting from a null model and ending with full interaction models. For all age comparisons, age was treated as an ordinal rather than a continuous variable to account for discrete timepoints of measurements and to allow for post-hoc pairwise comparisons between ages. All models had the individual animals as random effect. We used chi-square tests, Akaike Information Criterion (AIC), and Bayes Information Criterion (BIC) as criteria for model comparison. For models with significant main effects, we then used estimated marginal means for post-hoc comparisons. We used Bonferroni p-value correction to account for multiple comparisons. In all figures, we denote significance levels as follows: *, p(Bonf.) = 0.05; **, p(Bonf.) = 0.01; ***, p(Bonf.) = 0.001. All statistical testing was carried out in R 4.2.1 (Team, 2024) We used the lme4 package (Bates et al., 2015) for mixed effects modelling and emmeans package (Lenth, 2024) for post-hoc comparisons.

eLife Assessment

Zebra finches are a prominent model system for vocal learning and auditory system function, yet little is known about the functional development of the auditory system. Here, the authors convincingly show that newly hatched zebra finches lack detectable auditory brainstem responses and that auditory neural signals emerge only days after hatching, challenging influential claims of prenatal acoustic communication in altricial birds. This important work clarifies the developmental timeline for auditory communication and highlights the value of neuroscientific methods for validating and complementing behavioral ecological studies of animal perception.

Reviewer #1 (Public review):

This work by Antonnen et al. was triggered by claims of auditory-mediated effects on altricial avian embryos, which were published without any direct evidence that the relevant parental vocalizations were actually heard. I agree with Anttonen et al. that, based on the available evidence about avian auditory development, those claims are highly speculative and therefore necessitate more direct experimental verification.

Attonen et al. have embarked on a comprehensive series of experiments to:

(1) Better characterize acoustically the relevant parental vocalizations (heat whistles; in a separate preprint, not reviewed here)

(2) Characterize the auditory sensitivity of zebra finches at various stages of their posthatching development. Despite the long-standing importance of the zebra finch as a songbird model in neuroethology of learned vocalizations, the auditory development of the species has not been studied so far.

(3) Explore an alternative hypothesis of how the parental vocalizations might be perceived.

The principal method used here is the non-invasive recording of ABR (auditory brainstem response), a standard neurophysiological method in auditory research. The click-evoked ABR provides a quick and objective assessment of basic hearing sensitivity that does not require animal training. Weaknesses of the technique include its limited frequency specificity and low signal-to-noise ratio. The authors are experienced with ABR measurements and well aware of those issues. ABR responses in zebra finches are shown to gradually appear during the first week posthatching and to mature in subsequent weeks, consistent with the auditory development in other altricial bird species studied previously. When matching the acoustic properties of parental heat whistles and auditory sensitivities, hearing of the parental heat whistles by zebra finch hatchlings was convincingly excluded. Although not directly measured, this also convincingly extrapolates to zebra finch embryos. Finally, the authors tested the hypothesis that parental heat whistles could induce perceptible vibrations of the egg and thus stimulate the embryo via a different modality. The method used here was laser doppler vibrometry, an appropriate, state-of-the-art technique that the authors also have proven experience with. The induced vibrations were shown to be several orders of magnitude below known vibrotactile sensitivities in mammals and birds. Thus, although zebra finch vibrotactile thresholds were not obtained directly, the hypothesis of vibrotactile perception of parental heat whistles by zebra finch embryos could also be rejected convincingly.

In summary, even when considering some weaknesses of the techniques (which the authors are aware of), the conclusions of the paper are well supported: Auditory and/or vibration perception of parental heat whistles can be excluded as an explanation for previous reports of developmental programming for high ambient temperatures. As a constructive suggestion towards resolving the apparent paradox, the authors recommend repeating some of the crucial, previous playback experiments at lower sound levels that better match the natural parental vocalizations.

Reviewer #2 (Public review):

This study by Anttonen, Christensen-Dalsgaard, and Elemans describes the development of hearing thresholds in an altricial songbird species, the zebra finch. The results are very clear and along what might have been expected for altricial birds: at hatch (2 days post-hatch), the chicks are functionally deaf. Auditory evoked activity in the form of auditory brainstem responses (ABR) can start to be detected at 4 days post-hatch, but only at very loud sound levels. The study also shows that ABR response matures rapidly and reaches adult-like properties around 25 days post-hatch. The functional development of the auditory system is also frequency dependent, with a low-to-high frequency time course. All experiments are very well performed. The careful study throughout development and with the use of multiple time-points early in development is important to further ensure that the negative results found right after hatching are not the result of the experimental manipulation. The results themselves could be classified as somewhat descriptive, but, as the authors point out, they are particularly relevant and timely. Since 2016, there have been a series of studies published in high-profile journals that have presumably shown the importance of prenatal acoustic communication in altricial birds, mostly in zebra finches. This early acoustic communication would serve various adaptive functions. Although acoustic communication between embryos in the egg and parents has been shown in precocial birds (and crocodiles), finding an important function for prenatal communication in altricial birds came as a surprise. Unfortunately, none of those studies performed a careful assessment of the chicks' hearing abilities. This is done here, and the results are clear: zebra finches at 2 and 6 days post-hatch are functionally deaf. Since it is highly improbable that the hearing in the egg is more developed than at birth, one can only conclude that zebra finches in the egg (or at birth) cannot hear the heat whistles. The paper also ruled out the detection on egg vibrations as an alternative path. The prior literature will have to be corrected, or further studies conducted to solve the discrepancies. For this purpose, the "companion" paper on bioRxiv that studies the bioacoustical properties of heat calls from the same group will be particularly useful. Researchers from different groups will be able to precisely compare their stimuli.

Beyond the quality of the experiments, I also found that the paper was very well written. The introduction was particularly clear and complete (yet concise).

Weaknesses:

My only minor criticism is that the authors do not discuss potential differences between behavioral audiograms and ABRs. Optimally, one would need to repeat the work of Okanoya and Dooling with your setup and using the same calibration. The ~20dB difference might be real, or it might be due to SPL measured with different instruments, at different distances, etc. Either way, you could add a sentence in the discussion that states that even with the 20 dB difference in audiogram heat whistles would not be detected during the early days post-hatch. But adding a (novel) behavioral assay in young birds could further resolve the issue.

More Minor Points:

(1) As mentioned in the main text, the duration of pips (from pips to bursts) affects the effective bandwidth of the stimulus. I believe that the authors could give an estimate of this effective bandwidth, given what is known from bird auditory filters. I think that this estimate could be useful to compare to the effective bandwidth of the heat-call, which can now also be estimated.

(2) Figure 5b. Label the green and pink areas as song and heat-call spectrum. Also note that in the legend the authors say: "Green and red areas display the frequency windows related to the best hearing sensitivity of zebra finches and to heat calls, respectively". I don't think this is what they meant. I agree that 1-4 kHz is the best frequency sensitivity of zebra finches, but they probably meant green == "song frequency spectrum" and pink == "heat call spectrum". In either case, the figure and the legend need clarification.

(3) Figure 5c. Here also, I would change the song and heat-call labels to "song spectrum", "heat call spectrum". The authors would not want readers to think that they used song and heat calls in these experiments (maybe next time?). For the same reason, maybe in 5a you could add a cartoon of the oscillogram of a frequency sweep next to your speaker.

(4) Methods. In the description of the stimulus, the authors describe "5ms long tone bursts", but these are the tone pips in the main part of the manuscript. Use the same terms.

Reviewer #3 (Public review):

Summary

Following recent findings that exposure to natural sounds and anthropogenic noise before hatching affects development and fitness in an altricial songbird, this study attempts to estimate the hearing capacities of zebra finch nestlings and the perception of high frequencies in that species. It also tries to estimate whether airborne sound can make zebra finch eggs vibrate, although this is not relevant to the question.

Strength

That prenatal sounds can affect the development of altricial birds clearly challenges the long-held assumption that altricial avian embryos cannot hear. However, there is currently no data to support that expectation. Investigating the development of hearing in songbirds is therefore important, even though technically challenging. More broadly, there is accumulating evidence that some bird species use sounds beyond their known hearing range (especially towards high frequencies), which also calls for a reassessment of avian auditory perception.

Weaknesses

Rather than following validated protocols, the study presents many experimental flaws and two major methodological mistakes (see below), which invalidate all results on responses to frequency-specific tones in nestlings and those on vibration transmission to eggs, as well as largely underestimating hearing sensitivity. Accordingly, the study fails to detect a response in the majority of individuals tested with tones, including adults, and the results are overall inconsistent with previous studies in songbirds. The text throughout the preprint is also highly inaccurate, often presenting only part of the evidence or misrepresenting previous findings (both qualitatively and quantitatively; some examples are given below), which alters the conclusions.

Conclusion and impact

The conclusion from this study is not supported by the evidence. Even if the experiment had been performed correctly, there are well-recognised limitations and challenges of the method that likely explain the lack of response. The preprint fails to acknowledge that the method is well-known for largely underestimating hearing threshold (by 20-40dB in animals) and that it may not be suitable for a 1-gram hatchling. Unlike what is claimed throughout, including in the title, the failure to detect hearing sensitivity in this study does not invalidate all previous findings documenting the impacts of prenatal sound and noise on songbird development. The limitations of the approach and of this study are a much more parsimonious explanation. The incorrect results and interpretations, and the flawed representation of current knowledge, mean that this preprint regrettably creates more confusion than it advances the field.

Detailed assessment

For brevity, only some references are included below as examples, using, when possible, those cited in the preprint (DOI is provided otherwise). A full review of all the studies supporting the points below is beyond the scope of this assessment.

(A) Hearing experiment

The study uses the Auditory Brainstem Response (ABR), which measures minute electrical signals transmitted to the surface of the skull from the auditory nerve and nuclei in the brainstem. ABR is widely used, especially in humans, because it is non-invasive. However, ABR is also a lot less sensitive than other methods, and requires very specific experimental precautions to reliably detect a response, especially in extremely small animals and with high-frequency sounds, as here.

(1) Results on nestling frequency sensitivity are invalid, for failing to follow correct protocols:

The results on frequency testing in nestlings are invalid, since what might serve as a positive control did not work: in adults, no response was detected in a majority of individuals, at the core of their hearing range, with loud 95dB sounds (Figure S1), when testing frequency sensitivity with "tone burst".

This is mostly because the study used a stimulation duration 5 times larger than the norm. It used 25ms tone bursts, when all published avian studies (in altricial or precocial birds) used stimulation of 5ms or less (when using subdermal electrodes as here; e.g., cited: Brittan-Powell et al 2004; not cited: Brittan-Powell et al 2002 (doi: 10.1121/1.1494807), Henry & Lucas 2008 (doi: 10.1016/j.anbehav.2008.08.003)). Long stimulations do not make sense and are indeed known to interfere with the detection of an ABR response, especially at high frequencies, as, for example, explicitly tested and stated in Lauridsen et al 2021 (cited).

Adult response was then re-tested with a correct 5ms tone duration ("tone-pip"), which showed that, for the few individuals that responded to 25ms tones, thresholds were abnormally high (c.a. by 30dB; Figure 2C).

Yet, no nestlings were retested with a correct protocol. There is therefore no valid data to support any conclusion on nestling frequency hearing. Under these circumstances, the fact that some nestlings showed a response to 25ms tones from day 8 would argue against them having very low sensitivity to sound.

(2) Responses to clicks underestimate hearing onset by several days:

Without any valid nestling responses to tones (see # 1), establishing the onset of hearing is not possible based on responses to clicks only, since responses to clicks occur at least 4 days after responses to tones during development (Saunders et al, 1973). Here, 60% of 4-day-old individuals responding to clicks means most would have responded to tones at and before 2 days post-hatch, had the experiment been done correctly.

Responses to tones are indeed observed in other songbirds at 1day post-hatch (see #6).

In budgerigars, hearing onset occurs before 5 days post hatch, since responses to both clicks and tones were detectable at the first age tested at 5dph (Brittan-Powell et al, 2004).

(3) Experimental parameters chosen lower ABR detectability, specifically in younger birds:

Very fast stimulus repetition rate inhibits the ABR response, especially in young:

(a) The stimulus presentation rate (25 stim/ sec) is 6 times faster than zebra finch heat-calls, and 5 to 25 times faster than most previous studies in young birds (e.g., cited: Saunders et al 1973, 1974: 1 stim/sec or less; Katayama 1985: 3.3 clicks/sec; Brittan-Powell et al 2004: 4 stim/sec). Faster rates saturate the neurons and accordingly are known to decrease ABR amplitude and increase ABR latency, especially in younger animals with an immature nervous system. In birds, this occurs especially in the range from 5 to 30 stim/sec (e.g., cited: Saunder et al 1973, Brittan-Powell et al 2004). Values here with 25 rather than 1-4 stim/min are therefore underestimating true sensitivity.

(b) Averaging over only 400 measures is insufficient to reliably detect weak ABR signals:

The study uses 2 to 3 times fewer measures per stimulation type than the recommended value of 1,000 (e.g., Brittan-Powell et al 2002, 2024; Henry & Lucas 2008). This specifically affects the detection of weak signals, as in small hatchlings with tiny brains (adult zebra finches are 12-14g).

(c) Body temperature is not specified and strongly affects the ABR:

Controlling the body temperature of hatchlings of 1-4 grams (with a temperature probe under a 5mm-wide wing) would be very challenging. Low body temperature entirely eliminates the ABR, and even slight deviance from optimal temperature strongly increases wave latency and decreases wave amplitude (e.g., cited: Katayama 1985).

(d) Other essential information is missing on parameters known to affect the ABR:

This includes i) the weight of the animals, ii) whether and how the response signal was amplified and filtered, iii) how the automatised S/N>2 criteria compared to visual assessment for wave detection, and iv) what measures were taken to allow the correct placement of electrodes on hatchlings less than 5 grams.

(4) Results in adults largely underestimate sensitivity at high frequencies, and are not the correct reference point:

(a) Thresholds measured here at high frequencies for adults (using the correct stimulus duration, only done on adults) are 10-30dB higher than in all 3 other published ABR studies in adult zebra finches (cited: Zevin et al 2004; Amin et al 2007; not cited: Noirot et al 2011 (10.1121/1.3578452)), for both 4 and 6 kHz tone pips.

(b) The underlying assumption used throughout the preprint that hearing must be adult-like to be functional in nestlings does not make sense. Slower and smaller neural responses are characteristic of immature systems, but it does not mean signals are not being perceived.

(5) Failure to account for ABR underestimation leads to false conclusions:

(a) Whether the ABR method is suitable to assess hearing in very small hatchlings is unknown. No previous avian study has used ABR before 5 days post-hatch, and all have used larger bird species than the zebra finch.

(b) Even when performed correctly on large enough animals, the ABR systematically underestimates actual auditory sensitivity by 20-40 dB, especially at high frequencies, compared to behavioural responses (e.g., none cited: Brittan-Powell et al 2002, Henry & Lucas 2008, Noirot et al 2011). Against common practice, the preprint fails to account for this, leading to wrong interpretations. For example, in Figure 1G (comparing to heat call levels), actual hearing thresholds would be 30-40dB below those displayed. In addition, the "heat whistle" level displayed here (from the same authors) is 15dB lower than their second measure that they do not mention, and than measures obtained by others (unpublished data). When these two corrections are made - or even just the first one - the conclusion that heat-call sound levels are below the zebra finch hearing threshold does not hold.

(c) Rather than making appropriate corrections, the preprint uses a reference in humans (L180), where ABR is measured using a much more powerful method (multi-array EEG) than in animals, and from a larger brain. The shift of "10-20dB" obtained in humans is not applicable to animals.

(6) Results are inconsistent with previous findings in developing songbirds:

As expected from all of the above, results and conclusions in the preprint are inconsistent with findings in other songbirds, which, using other methods, show for example, auditory sensitivity in:

(a) zebra finch embryos, in response to song vs silence (not cited: Rivera et al 2018, doi: 10.1097/WNR.0000000000001187)

(b) flycatcher hatchlings at 2-3d post hatch (first age tested), across a wide range of frequencies (0.3 to 5kHz), at low to moderate sound levels (45-65dB) (cited: Aleksandrov and Dmitrieva 1992, not cited: Korneeva et al 2006 (10.1134/S0022093006060056)).

(c) songbird nestlings at 2-6d post hatch, which discriminate and behaviourally respond to relevant parental calls or even complex songs. This level of discrimination requires good hearing across frequencies (e.g., not cited: Korneeva et al 2006; Schroeder & Podos 2023 (doi: 10.1016/j.anbehav.2023.06.015)).

(d) zebra finch nestlings at 13d post-hatch, which show adult-like processing of songs in the auditory cortex (CNM) (Schroeder & Remage‐Healey 2021, doi: 10.1002/dneu.22802).

(e) zebra finch juveniles, which are able to perceive and learn song syllables at 5-7kHz (fundamental frequency) with very similar acoustic properties to heat calls, and also produced during inspiration (Goller & Daley 2001, doi: 10.1098/rspb.2001.1805).

NONE of these results - which contradict results and claims in the preprint - are mentioned. Instead, the preprint focuses on very slow-developing species (parrots and owls), which take 2-4 times longer than songbirds to fledge (cited: Brittan-Powell et al 2004; Köppl & Nickel 2007; Kraemer et al 2017).

(7) Results in figures are misreported in the text, and conclusions in the abstract and headers are not supported by the data:

For example:

(a) The data on Figure 1E shows that at 4 days old, 8 out of 13 nestlings (60%) responded to clicks, but the text says only 5/13 responded (L89). When 60% (4dph) and 90% (6dph) of individuals responded, the correct term would be that "most animals", rather than "some animals" responded (L89). Saying that ABR to loud sound appeared "in the majority only after one week" (L93) is also incorrect, given the data. It follows that the title of the paragraph is also erroneous.

(b) The hearing threshold is underestimated by 40dB at 6 and 8Kz on Fig 2C, not by "10-20dB" as reported in the text (L178).

(B) Egg vibration experiment

(8) Using airborne sound to vibrate eggs is biologically irrelevant:

The measurement of airborne sound levels to vibrate eggs misunderstands bone conduction hearing and is not biologically meaningful: zebra finch parents are in direct contact with the eggs when producing heat calls during incubation, not hovering in front of the nest. This misunderstanding affects all extrapolations from this study to findings in studies on prenatal communication.

(C) Misrepresentation of current knowledge

(9) Values from published papers are misreported, which reverses the conclusions:

Most critical examples:

(a) Preprint: "Zebra finch most sensitive hearing range of 1-to-4 kHz (Amin et al., 2007; Okanoya and Dooling, 1987; Yeh et al., 2023)" (L173).

Actual values in the studies cited are:

1-to-7kHz, in Amin et al 2007 (threshold [=50dB with ABR] is the same at 7kHz and 1KHz).

1-to-6 kHz, in Okanoya and Dooling (the threshold [=30dB with behaviour] is actually lower at 6kHz than at 1KHz).

1-to-7kHz, in Yeh et al (threshold [=35-38dB with behaviour] is the same at 7kHz and 1KHz).

Note that zebra finch nestlings' begging calls peaking at 6kHz (Elie & Theunissen 2015, doi: 10.1007/s10071-015-0933-6), would fall 2kHz above the parents' best hearing range if it were only up to 4kHz.

(b) The preprint incorrectly states throughout (e.g., L139, L163, L248) that heat-calls are 7-10kHz, when the actual value is 6-10kHz in the paper cited (Katsis et al, 2018).

(c) Using the correct values from these studies, and heat-calls at 45 dB SLP (as measured by others (unpublished data), or as measured by the authors themselves, but which is not reported here (Anttonen et a,l 2025), the correct conclusion is that heat calls fall within the known zebra finch hearing range.

(10) Published evidence towards high-frequency hearing, including in early development, is systematically omitted:

(a) Other studies showing birds use high frequencies above the known avian hearing range are ignored. This includes oilbirds (7-23kHz; Brinklov et al 2017; by 1 of the preprint authors, doi: 10.1098/rsos.170255) and hummingbirds (10-20kHz; Duque et al 2020, doi: 10.1126/sciadv.abb9393), and in a lesser extreme, zebra finches' inspiratory song syllables at 5-7kHz (Goller & Dalley, 2001).

(b) The discussion of anatomical development (L228-241) completely omits the well-known fact that the avian basilar papilla develops from high to low frequencies (i.e., base to apex), which - as many have pointed out - is opposite to the low-to-high development of sensitivity (e.g., cited: Cohen & Fermin 1978; Caus Capdevila et al 2021).

(c) High frequency hearing in songbirds at hatching is several orders of magnitude better than in chickens and ducks at the same age, even though songbirds are altricial (e.g., at 4kHz, flycatcher: 47dB, chicken-duck: 90dB; at 5kHz, flycatcher: 65dB, chicken-duck: 115dB; Korneeva et al 2006, Saunders et al 1974). That is because Galliformes are low-frequency specialists, according to both anatomical and ecological evidence, with calls peaking at 0.8 to 1.2kHz rather than 2-6kHz in songbirds. It is incorrect to conclude that altricial embryos cannot perceive high frequencies because low-frequency specialist precocial birds do not (L250;261).

The references used to support the statement on a very high threshold for precocial birds above 6kHz are also wrong (L250). Katayama 1985 did not test embryos, nor frequency tones. Neither of these two references tested ducks.

(11) Incorrect statements do not reflect findings from the references cited

For example:

(a) "in altricial bird species hearing typically starts after hatching" (L12, in abstract), "with little to no functional hearing during embryonic stages (Woolley, 2017)." (L33).

There is no evidence, in any species, to support these statements. This is only a - commonly repeated - assumption, not actually based on any data. On the contrary, the extremely limited evidence to date shows the opposite, with zebra finch embryos showing ZENK activation in the auditory cortex in response to song playback (Rivera et al, 2018, not cited).

The book chapter cited (Woolley 2017) acknowledges this lack of evidence, and, in the context of song learning, provides as only references (prior to 2018), 2 studies showing that songbirds do not develop a normal song if the song tutor is removed before 10d post-hatch. That nestlings cannot memorise (to later reproduce) complex signals heard before d10 does not mean that they are deaf to any sound before day 10.

Studies showing hearing in young songbird nestlings (see point 6 above) also contradict these statements.

(b) "Zebra finch embryos supposedly are epigenetically guided to adapt to high temperatures by their parents high-frequency "heat calls" " (L36 and L135).

This is an extremely vague and meaningless description of these results, which cannot be assessed by readers, even though these results are presented as a major justification for the present study. Rather than giving an interpretation of what "supposedly" may occur, it would be appropriate to simply synthesize the empirical evidence provided in these papers. They showed that embryonic exposure to heat-calls, as opposed to control contact calls, alters a suite of physiological and behavioural traits in nestlings, including how growth and cellular physiology respond to high temperatures. This also leads to carry-over effects on song learning and reproductive fitness in adulthood.

(c) "The acoustic communication in precocial mallard ducks depends specifically on the low-frequency auditory sensitivity of the embryo (Gottlieb, 1975)" (L253)

The study cited (Gottlieb, 1975) demonstrates exactly the opposite of this statement: it shows that duckling embryos, not only perceive high frequency sounds (relative to the species frequency range), but also NEED this exposure to display normal audition and behaviour post-hatch. Specifically, it shows that duckling embryos deprived of exposure to their own high-frequency calls (at 2 kHz), failed to identify maternal calls post-hatch because of their abnormal insensitivity to higher frequencies, which was later confirmed by directly testing their auditory perception of tones (Dimitrieva & Gottlieb, 1994).

(12) Considering all of the mistakes and distortions highlighted above, it would be very premature to conclude, based on these results and statements, that altricial avian embryos are not sensitive to sound. This study provides no actual scientific ground to support this conclusion.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

This work by Antonnen et al. was triggered by claims of auditory-mediated effects on altricial avian embryos, which were published without any direct evidence that the relevant parental vocalizations were actually heard. I agree with Anttonen et al. that, based on the available evidence about avian auditory development, those claims are highly speculative and therefore necessitate more direct experimental verification.

Attonen et al. have embarked on a comprehensive series of experiments to:

(1) Better characterize acoustically the relevant parental vocalizations (heat whistles; in a separate preprint, not reviewed here)

(2) Characterize the auditory sensitivity of zebra finches at various stages of their posthatching development. Despite the long-standing importance of the zebra finch as a songbird model in neuroethology of learned vocalizations, the auditory development of the species has not been studied so far.

(3) Explore an alternative hypothesis of how the parental vocalizations might be perceived.

The principal method used here is the non-invasive recording of ABR (auditory brainstem response), a standard neurophysiological method in auditory research. The click-evoked ABR provides a quick and objective assessment of basic hearing sensitivity that does not require animal training. Weaknesses of the technique include its limited frequency specificity and low signal-to-noise ratio. The authors are experienced with ABR measurements and well aware of those issues. ABR responses in zebra finches are shown to gradually appear during the first week posthatching and to mature in subsequent weeks, consistent with the auditory development in other altricial bird species studied previously. When matching the acoustic properties of parental heat whistles and auditory sensitivities, hearing of the parental heat whistles by zebra finch hatchlings was convincingly excluded. Although not directly measured, this also convincingly extrapolates to zebra finch embryos. Finally, the authors tested the hypothesis that parental heat whistles could induce perceptible vibrations of the egg and thus stimulate the embryo via a different modality. The method used here was laser doppler vibrometry, an appropriate, state-of-the-art technique that the authors also have proven experience with. The induced vibrations were shown to be several orders of magnitude below known vibrotactile sensitivities in mammals and birds. Thus, although zebra finch vibrotactile thresholds were not obtained directly, the hypothesis of vibrotactile perception of parental heat whistles by zebra finch embryos could also be rejected convincingly.

In summary, even when considering some weaknesses of the techniques (which the authors are aware of), the conclusions of the paper are well supported: Auditory and/or vibration perception of parental heat whistles can be excluded as an explanation for previous reports of developmental programming for high ambient temperatures. As a constructive suggestion towards resolving the apparent paradox, the authors recommend repeating some of the crucial, previous playback experiments at lower sound levels that better match the natural parental vocalizations.

(R1. A1) We thank the reviewer for their time and effort to thoroughly review our paper and for the positive comments on our manuscript. In the revised manuscript we have addressed the concerns that you have raised in the Joint recommendations above (Pages 1-4).

Reviewer #2 (Public review):

This study by Anttonen, Christensen-Dalsgaard, and Elemans describes the development of hearing thresholds in an altricial songbird species, the zebra finch. The results are very clear and along what might have been expected for altricial birds: at hatch (2 days post-hatch), the chicks are functionally deaf. Auditory evoked activity in the form of auditory brainstem responses (ABR) can start to be detected at 4 days post-hatch, but only at very loud sound levels. The study also shows that ABR response matures rapidly and reaches adult-like properties around 25 days post-hatch. The functional development of the auditory system is also frequency dependent, with a low-to-high frequency time course. All experiments are very well performed. The careful study throughout development and with the use of multiple time-points early in development is important to further ensure that the negative results found right after hatching are not the result of the experimental manipulation. The results themselves could be classified as somewhat descriptive, but, as the authors point out, they are particularly relevant and timely. Since 2016, there have been a series of studies published in high-profile journals that have presumably shown the importance of prenatal acoustic communication in altricial birds, mostly in zebra finches. This early acoustic communication would serve various adaptive functions. Although acoustic communication between embryos in the egg and parents has been shown in precocial birds (and crocodiles), finding an important function for prenatal communication in altricial birds came as a surprise. Unfortunately, none of those studies performed a careful assessment of the chicks' hearing abilities. This is done here, and the results are clear: zebra finches at 2 and 6 days post-hatch are functionally deaf. Since it is highly improbable that the hearing in the egg is more developed than at birth, one can only conclude that zebra finches in the egg (or at birth) cannot hear the heat whistles. The paper also ruled out the detection on egg vibrations as an alternative path. The prior literature will have to be corrected, or further studies conducted to solve the discrepancies. For this purpose, the "companion" paper on bioRxiv that studies the bioacoustical properties of heat calls from the same group will be particularly useful. Researchers from different groups will be able to precisely compare their stimuli.

Beyond the quality of the experiments, I also found that the paper was very well written. The introduction was particularly clear and complete (yet concise).

Weaknesses:

My only minor criticism is that the authors do not discuss potential differences between behavioral audiograms and ABRs. Optimally, one would need to repeat the work of Okanoya and Dooling with your setup and using the same calibration. The ~20dB difference might be real, or it might be due to SPL measured with different instruments, at different distances, etc. Either way, you could add a sentence in the discussion that states that even with the 20 dB difference in audiogram heat whistles would not be detected during the early days post-hatch. But adding a (novel) behavioral assay in young birds could further resolve the issue.

(R2. A0) We thank the reviewer for their time and effort to thoroughly review our paper, and for the positive comments on our manuscript.

In our revision, we have added a new figure (Fig 5) and three new paragraphs (Lines 387-422) in the discussion to compare all published ABR and behavioral audiograms and the differences between and among these datasets. Our adult data is consistent with the reported findings from four other labs despite differences in stimulus design, setups and genetic background of the animals, providing strong support for our findings.

Furthermore we have more clearly presented our argument why we think juvenile and embryos cannot detect heat whistles. For clarification, we have added a new figure (Fig 4) and four new paragraphs (Lines 271-355) in the discussion.

We agree with the reviewer that more data is needed on the development of hearing in songbirds and zebra finches especially; both anatomical data and functional, such as innervation. We emphasize this need in our discussion (Lines 352-355 and Lines 405-411).

More Minor Points:

(1) As mentioned in the main text, the duration of pips (from pips to bursts) affects the effective bandwidth of the stimulus. I believe that the authors could give an estimate of this effective bandwidth, given what is known from bird auditory filters. I think that this estimate could be useful to compare to the effective bandwidth of the heat-call, which can now also be estimated.

(R2. A1) Please see answer A3.4 under Joint Recommendations.

(2) Figure 5b. Label the green and pink areas as song and heat-call spectrum. Also note that in the legend the authors say: "Green and red areas display the frequency windows related to the best hearing sensitivity of zebra finches and to heat calls, respectively". I don't think this is what they meant. I agree that 1-4 kHz is the best frequency sensitivity of zebra finches, but they probably meant green == "song frequency spectrum" and pink == "heat call spectrum". In either case, the figure and the legend need clarification.

(R2. A2) We thank the reviewer for pointing out these issues. We have changed the figure and legend accordingly. In the meantime, we published a paper measuring the in vivo source levels of the heat whistles (Anttonen et al., Current Biology 2025), and we have carefully gone through this manuscript to incorporate those findings and adjust the text accordingly. We have therefore adjusted the analysis in Fig 3B to correct for the narrower frequency distribution of the heat whistles.

(3) Figure 5c. Here also, I would change the song and heat-call labels to "song spectrum", "heat call spectrum". The authors would not want readers to think that they used song and heat calls in these experiments (maybe next time?). For the same reason, maybe in 5a you could add a cartoon of the oscillogram of a frequency sweep next to your speaker.

(R2. A3) We thank the reviewer for pointing out these issues We mention the frequency sweep in the legend for panel A, but decided against including this in the figure to prevent too much clutter. We have changed the figure = legend to (new text underlined):

Legend Fig 3. “A Setup used to measure sound-induced vibrations of eggs. A 94 dB, 0.25-to-10 kHz frequency sweep was played at the eggs to determine the vibration transfer function.”

(4) Methods. In the description of the stimulus, the authors describe "5ms long tone bursts", but these are the tone pips in the main part of the manuscript. Use the same terms.

(R2. A4) Thank you for catching this, we have changed this into “5ms long tone pips".

Reviewer #3 (Public review):

Summary

Following recent findings that exposure to natural sounds and anthropogenic noise before hatching affects development and fitness in an altricial songbird, this study attempts to estimate the hearing capacities of zebra finch nestlings and the perception of high frequencies in that species. It also tries to estimate whether airborne sound can make zebra finch eggs vibrate, although this is not relevant to the question.

Strength

That prenatal sounds can affect the development of altricial birds clearly challenges the long-held assumption that altricial avian embryos cannot hear. However, there is currently no data to support that expectation. Investigating the development of hearing in songbirds is therefore important, even though technically challenging. More broadly, there is accumulating evidence that some bird species use sounds beyond their known hearing range (especially towards high frequencies), which also calls for a reassessment of avian auditory perception.

Weaknesses

Rather than following validated protocols, the study presents many experimental flaws and two major methodological mistakes (see below), which invalidate all results on responses to frequencyspecific tones in nestlings and those on vibration transmission to eggs, as well as largely underestimating hearing sensitivity. Accordingly, the study fails to detect a response in the majority of individuals tested with tones, including adults, and the results are overall inconsistent with previous studies in songbirds. The text throughout the preprint is also highly inaccurate, often presenting only part of the evidence or misrepresenting previous findings (both qualitatively and quantitatively; some examples are given below), which alters the conclusions.

Conclusion and impact

The conclusion from this study is not supported by the evidence. Even if the experiment had been performed correctly, there are well-recognised limitations and challenges of the method that likely explain the lack of response. The preprint fails to acknowledge that the method is well-known for largely underestimating hearing threshold (by 20-40dB in animals) and that it may not be suitable for a 1-gram hatchling. Unlike what is claimed throughout, including in the title, the failure to detect hearing sensitivity in this study does not invalidate all previous findings documenting the impacts of prenatal sound and noise on songbird development. The limitations of the approach and of this study are a much more parsimonious explanation. The incorrect results and interpretations, and the flawed representation of current knowledge, mean that this preprint regrettably creates more confusion than it advances the field.

(R3. A0) We thank the reviewer for their detailed and critical assessment. We agree that establishing auditory sensitivity in very young altricial birds is technically challenging and that careful interpretation of ABR data is essential. We also appreciate the reviewer’s recognition of the importance of obtaining direct physiological data on auditory development.

However, we respectfully disagree with the reviewer’s central claim that our methodology is flawed or that our conclusions are unsupported. Many of the concerns raised reflect misunderstandings of ABR methodology, selective interpretation of the literature, or assumptions that are not supported by empirical evidence. Below, we address the main points in turn.

As also stated in our Provisional response, the reviewer’s critique can be distilled into four main arguments:

(1) ABR cannot be reliably measured in very small animals.

(2) Our stimulus design (especially 25 ms tone bursts) invalidates frequency-specific results.

(3) ABR thresholds should be corrected to behavioral thresholds, which would alter conclusions.

(4) Our findings are inconsistent with prior studies in songbirds.

We address each of these below before responding point-by-point.

(1) Suitability of ABR in small animals.

Reviewer claim: ABR may not be suitable for very small hatchlings.

This claim is not supported by existing evidence. ABR measures summed neural activity, and signal amplitude depends in part on the distance between neural tissue and recording electrodes. In smaller animals, this distance is reduced, which can increase signal amplitude and improve signal-to-noise ratio.

Consistent with this, ABR has been successfully recorded in animals substantially smaller than zebra finch hatchlings, including zebrafish (Jørgensen et al., 2012), 10 mm froglets (Goutte et al., 2017) and 5 mm salamanders (Capshaw et al., 2020). It is in fact much more surprising the technique still provides robust signals even in extremely large animals such as Minke whales, where the distance between electrodes and brain is on the decimeter scale (Houser et al., 2024). We have extensive experience of recording ABRs in such small systems.

Thus, there is no principled reason why ABR would be an invalid method to study auditory sensitivity in zebra finch hatchlings.

(2) Stimulus design and tone duration

Reviewer claim: Use of 25 ms tone bursts invalidates frequency-specific results.

We agree that stimulus duration affects frequency specificity and ABR detectability. However, the reviewer’s assertion that there is a single “correct protocol” (≤5 ms) is inaccurate. In avian ABR studies, stimulus duration varies depending on experimental goals.

Our choice of 25 ms tone bursts was intentional and necessary to accurately represent low frequencies (down to 250 Hz), ensuring sufficient cycles per stimulus in the plateau segment of 15 ms and minimizing spectral splatter (see auditory brainstem response design considerations discussed in Lauridsen et al., 2021) and our responses below.

Key clarifications:

a) Click-evoked ABRs form the basis of our conclusions about onset of hearing, not tone bursts.

b) Tone bursts were used primarily to assess frequency-dependent maturation, not detect earliest sensitivity.

c) We explicitly demonstrate that:

- 25 ms bursts yield higher thresholds (lower sensitivity)

- 5 ms pips yield lower thresholds and align with published ABR audiograms

We have now:

- Further clarified the rationale of stimulus design in the methods (Line 520-530) and added a section in the discussion (Lines 397-411).

- Included additional comparison between burst and pip datasets (Lines 387-395).

- Clarified that conclusions about early hearing do not depend on tone-burst data (Fig 4 and Lines 271-294).

(3) ABR vs behavioral thresholds

Reviewer claim: Failure to correct ABR thresholds (20–40 dB) invalidates conclusions.

We agree that ABR thresholds typically overestimate behavioral thresholds. However, we disagree that this invalidates our conclusions.

Importantly:

a) We do not replace measured ABR data with corrected values, as this would be methodologically inappropriate.

b) Instead, we:

- Present measured ABR thresholds transparently (Fig 1-3)

- Compare them directly to published behavioral audiograms (Fig 5)

- Explicitly discuss the expected offset (Lines 413-422)

In the revised manuscript we:

a) Add a new figure (Fig. 5) compiling all published ABR and behavioral audiograms

b) Show that:

- ABR and behavioral audiograms have similar shapes (Fig 5, new discussion Lines 387-422)

- Offsets are typically ~20 dB (Line 413-422)

Crucially, even under conservative corrections:

- Early hatchlings remain far less sensitive than adults (>54 dB SPL) to clicks.

- Heat whistle levels remain at or below detection limits even in adults (new figure Fig 4)

- The developmental gap (>50 dB between adults and 2 DPH hatchlings) remains decisive.

Thus, incorporating ABR–behavioral differences does not change the central conclusion.

(4) Consistency with prior literature in developing songbirds.

Reviewer claim: Results contradict previous studies in developing songbirds.

We respectfully disagree. The cited studies fall into three categories:

(1) Behavioral studies (e.g., alarm-call responses)

(2) Gene expression studies (e.g., ZENK activation)

(3) Different species with different developmental trajectories

None of these directly measure auditory sensitivity thresholds in zebra finch embryos or hatchlings.

We emphasize:

- Behavioral responses do not provide threshold measurements

- Neural activation (e.g., ZENK) does not demonstrate functional perception thresholds

- Cross-species comparisons must consider differences in developmental timing.

We have now expanded the Discussion to explicitly address these studies and clarify how they relate to our findings (Lines 357-367). Our data are consistent with what is known about the physiology of auditory development in all birds studied so far.

(5) Final statement

We have revised the manuscript extensively to:

- Clarify methodology and experimental design

- Expand discussion of ABR limitations

- Incorporate additional literature and comparisons

- Correct inconsistencies in reporting

We maintain that our central conclusions—that early zebra finch hatchlings lack detectable auditory brainstem responses and are unlikely to perceive parental heat calls at natural levels—is robust and supported by the data. We go into more detail in our point-by-point rebuttal below.

Detailed assessment

For brevity, only some references are included below as examples, using, when possible, those cited in the preprint (DOI is provided otherwise). A full review of all the studies supporting the points below is beyond the scope of this assessment.

(A) Hearing experiment

The study uses the Auditory Brainstem Response (ABR), which measures minute electrical signals transmitted to the surface of the skull from the auditory nerve and nuclei in the brainstem. ABR is widely used, especially in humans, because it is non-invasive. However, ABR is also a lot less sensitive than other methods, and requires very specific experimental precautions to reliably detect a response, especially in extremely small animals and with high-frequency sounds, as here.

(1) Results on nestling frequency sensitivity are invalid, for failing to follow correct protocols:

(R3. A1.1) We disagree that our protocol is invalid. There is no universal ABR protocol standard in birds. Our approach is consistent with established principles of stimulus design and is validated by:

- Robust click-evoked responses

- Consistent developmental trajectories

- Agreement between pip-based ABR and behavioral audiograms

We now clarify this explicitly. See response R3. A0 (Stimulus design and tone duration).

The results on frequency testing in nestlings are invalid, since what might serve as a positive control did not work: in adults, no response was detected in a majority of individuals, at the core of their hearing range, with loud 95dB sounds (Figure S1), when testing frequency sensitivity with "tone burst".

This is mostly because the study used a stimulation duration 5 times larger than the norm. It used 25ms tone bursts, when all published avian studies (in altricial or precocial birds) used stimulation of 5ms or less (when using subdermal electrodes as here; e.g., cited: Brittan-Powell et al 2004; not cited: Brittan-Powell et al 2002 (doi: 10.1121/1.1494807), Henry & Lucas 2008 (doi: 10.1016/j.anbehav.2008.08.003)). Long stimulations do not make sense and are indeed known to interfere with the detection of an ABR response, especially at high frequencies, as, for example, explicitly tested and stated in Lauridsen et al 2021 (cited).

(R3. A1.2) ABR with long-duration stimuli were shown previously to work perfectly well in birds and do not interfere with the detection of an ABR response. Longer stimuli have been used, e.g. in the following bird papers:

- Amin et al., J Neurophysiol 2007 (cited) on zebra finch ABR: 20 ms tone bursts

- Korneeva et al. (2006), evoked responses from field L in flycatcher (cited by the reviewer): 20 ms

- Saunders et al. (1973) (cited): 60 ms tone bursts.

- Larsen ON, Wahlberg M, Christensen-Dalsgaard (2020) Amphibious hearing in a diving bird, the great cormorant (Phalacrocorax carbo sinensis), J Exp Biol, doi:10.1242/jeb.217265: 25 ms tone bursts

Human ABR has been measured even with long-duration speech signals (duration 40 ms and longer). See for example Binkhamis et al., Ear and Hearing 40: 659-670, 2019.

Furthermore, the reviewer unfortunately misunderstood some aspects in Lauridsen et al., which tested a specifical method exploiting neural phase locking, and showed that this method has a low-frequency bias because neural phase locking decreases at high frequencies.

Also, Lauridsen et al clearly show the reason for using 25 ms bursts: because we aimed to measure frequencies down to 250 Hz, we need to have a sufficient number of cycles (3) in the plateau segment to represent the frequency adequately (Lauridsen et al. fig 1). A 15 ms plateau contains 3 cycles, plus 5 ms rise/fall time (1 cycle) equals 25 ms. We have clarified this in our methods (L520-525) and added a section in the discussion (L397-411).

Thus, long-duration stimulations make sense and have been used before successfully. See also response R3.A0 (Stimulus design and tone duration).

Adult response was then re-tested with a correct 5ms tone duration ("tone-pip"), which showed that, for the few individuals that responded to 25ms tones, thresholds were abnormally high (c.a. by 30dB; Figure 2C).

Yet, no nestlings were retested with a correct protocol. There is therefore no valid data to support any conclusion on nestling frequency hearing. Under these circumstances, the fact that some nestlings showed a response to 25ms tones from day 8 would argue against them having very low sensitivity to sound.

(R3. A1.3) Please see answer A3.1 under Joint Recommendations and R3.A0 (Stimulus design and tone duration).

(2) Responses to clicks underestimate hearing onset by several days:

Without any valid nestling responses to tones (see # 1), establishing the onset of hearing is not possible based on responses to clicks only, since responses to clicks occur at least 4 days after responses to tones during development (Saunders et al, 1973). Here, 60% of 4-day-old individuals responding to clicks means most would have responded to tones at and before 2 days post-hatch, had the experiment been done correctly.

(R3. A2.1) We disagree that clicks necessarily underestimate onset.

Clicks are broadband stimuli that:

- Recruit large neural populations

- Are commonly used to detect early auditory responses

The cited delay between tone and click responses reflects stimulus energy differences, not an inherent limitation of clicks. We have clarified this in the revision and softened language to refer to “no detectable ABR response” rather than absolute deafness.

The report that Saunders could only see responses to clicks later than to tones only reflects that he used click amplitudes that were insufficiently high. The ABR responses reported were to extremely intense tones (110 dB SPL) of long duration (60 ms).

If Saunders had used clicks (duration 60 µs) with comparable sound energy, they would have had been very difficult to produce. He should have used clicks with an amplitude 1000 times (60 dB) higher than the tones to produce the same sound energy. This would have been clicks at 170 dB SPL, equivalent to the sound at the mouth of a medium-sized military cannon. Applying this pressure would not be a recommendable method for hearing assessment, but instead lead to irreversible hearing damage.

In budgerigars, hearing onset occurs before 5 days post hatch, since responses to both clicks and tones were detectable at the first age tested at 5dph (Brittan-Powell et al, 2004).

(R3. A2.2) This is not how we interpret the cited paper. They state that ‘Responses were first obtained from 1-week-old at high stimulation, and their click responses (Fig. 1) show no wave 1 peak at 6 days post-hatch. Also, their conclusion (p 3101) states that ‘budgerigars probably cannot hear at hatching’.

(3) Experimental parameters chosen lower ABR detectability, specifically in younger birds: Very fast stimulus repetition rate inhibits the ABR response, especially in young:

(a) The stimulus presentation rate (25 stim/ sec) is 6 times faster than zebra finch heat-calls, and 5 to 25 times faster than most previous studies in young birds (e.g., cited: Saunders et al 1973, 1974: 1 stim/sec or less; Katayama 1985: 3.3 clicks/sec; Brittan-Powell et al 2004: 4 stim/sec).

Faster rates saturate the neurons and accordingly are known to decrease ABR amplitude and increase ABR latency, especially in younger animals with an immature nervous system.

In birds, this occurs especially in the range from 5 to 30 stim/sec (e.g., cited: Saunder et al 1973, Brittan-Powell et al 2004). Values here with 25 rather than 1-4 stim/min are therefore underestimating true sensitivity.

(R3. A3a) Please see answer A3.3 under Joint Recommendations.

(b) Averaging over only 400 measures is insufficient to reliably detect weak ABR signals: The study uses 2 to 3 times fewer measures per stimulation type than the recommended value of 1,000 (e.g., Brittan-Powell et al 2002, 2024; Henry & Lucas 2008). This specifically affects the detection of weak signals, as in small hatchlings with tiny brains (adult zebra finches are 12-14g).

(R3. A3b) Please see answer A3.2 under Joint Recommendations.

(c) Body temperature is not specified and strongly affects the ABR:

Controlling the body temperature of hatchlings of 1-4 grams (with a temperature probe under a 5mm-wide wing) would be very challenging. Low body temperature entirely eliminates the ABR, and even slight deviance from optimal temperature strongly increases wave latency and decreases wave amplitude (e.g., cited: Katayama 1985).

(R3. A3c) Please see answer A2 under Joint Recommendations.

(d) Other essential information is missing on parameters known to affect the ABR: This includes i) the weight of the animals,

(R3. A3d-i) These important details have now been added in Table S6.

(ii) whether and how the response signal was amplified and filtered,

(R3. A3d-ii) Signal was amplified 500 times (74 dB). We have included these important details to the methods (Line 508).

(iii) how the automatised S/N>2 criteria compared to visual assessment for wave detection,

(R3. A3d-iii) There is no universally accepted/fitting protocol for performing ABR recordings in various animals, we decided to perform both visual and automated criteria detection of thresholds. The automated criterion in our experience is more strict approach than visual detection of thresholds, because using an automated criteria for threshold detection will remove potential experimenter bias from the results. We added this in our methods (Lines 583-585).

(iv) what measures were taken to allow the correct placement of electrodes on hatchlings less than 5 grams.

(R3. A3d-iv) We have placed electrodes in much smaller animals than 5 grams, and the common landmarks (ear opening, midline of skull) could easily be identified in the hatchlings.

(4) Results in adults largely underestimate sensitivity at high frequencies, and are not the correct reference point:

(a) Thresholds measured here at high frequencies for adults (using the correct stimulus duration, only done on adults) are 10-30dB higher than in all 3 other published ABR studies in adult zebra finches (cited: Zevin et al 2004; Amin et al 2007; not cited: Noirot et al 2011 (10.1121/1.3578452)), for both 4 and 6 kHz tone pips.

(b) The underlying assumption used throughout the preprint that hearing must be adult-like to be functional in nestlings does not make sense. Slower and smaller neural responses are characteristic of immature systems, but it does not mean signals are not being perceived.

(R3. A4) We acknowledge variation across studies and now include a comprehensive comparison of all zebra finch audiograms (new Fig. 5 and discussion Lines 387-422).

Importantly:

- Our pip-based audiogram aligns with previous ABR studies

- Differences in high-frequency sensitivity likely reflect methodological variation or population differences.

Our conclusions rely on relative developmental changes, not absolute thresholds.

(5) Failure to account for ABR underestimation leads to false conclusions:

(a) Whether the ABR method is suitable to assess hearing in very small hatchlings is unknown. No previous avian study has used ABR before 5 days post-hatch, and all have used larger bird species than the zebra finch.

(R3. A5a) As stated above (R3.A0), small animals should give better signals, and we have been able to measure ABR in much smaller animals previously.

(b) Even when performed correctly on large enough animals, the ABR systematically underestimates actual auditory sensitivity by 20-40 dB, especially at high frequencies, compared to behavioural responses (e.g., none cited: Brittan-Powell et al 2002, Henry & Lucas 2008, Noirot et al 2011). Against common practice, the preprint fails to account for this, leading to wrong interpretations.

(R3. A5a) See our answer to R3.A0 above.

For example, in Figure 1G (comparing to heat call levels), actual hearing thresholds would be 3040dB below those displayed. In addition, the "heat whistle" level displayed here (from the same authors) is 15dB lower than their second measure that they do not mention, and than measures obtained by others (unpublished data). When these two corrections are made - or even just the first one - the conclusion that heat-call sound levels are below the zebra finch hearing threshold does not hold.

(R3. A5a) Our conclusion that heat whistles are unlikely to be perceived does not rely on a single dataset or method, but on the convergence of three independent constraints: (i) signal amplitude, (ii) adult auditory sensitivity, and (iii) developmental immaturity of the auditory system.

First, heat whistles are low-amplitude signals. Our in vivo measurements show levels of ~33 dB re 20 µPa at 10 cm and ~14 dB at 1 m (Anttonen et al., 2025, Curr Biol). Even allowing for uncertainty in near-field estimation, a conservative upper bound at very close range (<5 cm) is ~40 dB SPL.

Second, the most sensitive available measure—behavioral audiograms—places adult zebra finch thresholds at ~40 dB SPL at ~6 kHz (Okanoya and Dooling, 1987, J Comp Psychol), increasing steeply toward higher frequencies. Thus, even under optimal conditions, heat whistles fall at or below the detection threshold of adults, and only potentially at very close range.

Third, auditory sensitivity in early development is substantially reduced. Our ABR data show a ≥40–60 dB decrease in sensitivity in hatchlings relative to adults for click stimuli, which provide the most favorable conditions for eliciting responses. Because frequency-specific sensitivity develops later, thresholds at 6–8 kHz are expected to be even higher in hatchlings and embryos.

Taken together, these constraints define a narrow and unfavorable detection window: a low-amplitude, high-frequency signal positioned at the edge of adult sensitivity, combined with a large developmental decrease in auditory sensitivity. Under these conditions, it is unlikely that heat whistles are detectable by hatchlings or embryos.

Importantly, this conclusion does not depend on precise correction factors between ABR and behavioral thresholds. Even when considering the most sensitive behavioral data and conservative estimates of sound level, the signal remains at or below the limits of detection in adults, and far below expected sensitivity in early developmental stages.

We have included this argument more clearly in our discussion (Line 271-367), illustrated by new Fig. 5.

(c) Rather than making appropriate corrections, the preprint uses a reference in humans (L180), where ABR is measured using a much more powerful method (multi-array EEG) than in animals, and from a larger brain. The shift of "10-20dB" obtained in humans is not applicable to animals.

(R3. A5c) Again our conclusions rely on relative developmental changes, not absolute thresholds. The clinical practice in humans to measure ABR is with 4 electrodes, not a multi-electrode EEG array.

Animal studies where ABR audiograms have been compared directly to psychophysical audiograms show differences of around 20 dB. For example in Brittain powell et al 2002, audiogram comparisons between behavioral and ABR in budgerigars were made within the same lab, same animal population and by the same people, leading to 20 dB difference. Our discussion includes a new paragraph on this topic (Lines 413-422).

(6) Results are inconsistent with previous findings in developing songbirds:

(R3. A6) We now explicitly discuss all cited studies. Key points:

- Early behavioral responses do not imply high-frequency sensitivity

- Studies in other species do not directly translate to zebra finches

- None of the cited work provides direct measures of auditory thresholds in embryos

As expected from all of the above, results and conclusions in the preprint are inconsistent with findings in other songbirds, which, using other methods, show for example, auditory sensitivity in: a) zebra finch embryos, in response to song vs silence (not cited: Rivera et al 2018, doi: 10.1097/WNR.0000000000001187)

(R3. A6a) We thank the reviewer for pointing out this study. We agree that the question of auditory responsiveness in embryos is important, and that a range of approaches have been used to address it. However, the study cited (Rivera et al., 2018) does not directly measure auditory sensitivity, but instead infers auditory processing from differences between treatment groups exposed to different acoustic conditions. As such, it is not directly comparable to physiological measures of hearing sensitivity, such as ABR or behavioral thresholds.

In addition, interpretation of these results is complicated by limited characterization of the acoustic environment and differences in experimental handling between groups, which may introduce confounding factors unrelated to auditory perception. Given these considerations, and because our study focuses specifically on quantifying auditory sensitivity using established physiological methods, we have chosen not to include a detailed discussion of this work.

(b) flycatcher hatchlings at 2-3d post hatch (first age tested), across a wide range of frequencies (0.3 to 5kHz), at low to moderate sound levels (45-65dB) (cited: Aleksandrov and Dmitrieva 1992, not cited: Korneeva et al 2006 (10.1134/S0022093006060056)).

(R3. A6b) Korneeva et al. (2006) and Aleksandrov and Dmitrieva (1992) report evoked responses in very young flycatcher hatchlings across a broad frequency range. Notably, these measurements were obtained using more invasive recording approaches (e.g., implanted electrodes in Field L) in unanesthetized birds, which are known to yield lower thresholds compared to far-field ABR recordings under anesthesia. These methodological differences likely account for part of the higher sensitivity reported.

Importantly, even in flycatchers, auditory sensitivity shows substantial postnatal improvement: thresholds decrease by up to ~40 dB over the first days after hatching, and the upper frequency limit expands from ~4 to ~7 kHz. Thus, while absolute sensitivity may differ across species and methods, the overall developmental trajectory—gradual improvement in sensitivity and progressive extension toward higher frequencies—is consistent with our findings and with broader patterns reported in songbirds.

We have added this paper in our discussion (Lines 362-365).

(c) songbird nestlings at 2-6d post hatch, which discriminate and behaviourally respond to relevant parental calls or even complex songs. This level of discrimination requires good hearing across frequencies (e.g., not cited: Korneeva et al 2006; Schroeder & Podos 2023 (doi: 10.1016/j.anbehav.2023.06.015)).

(R3. A6c) The species mentioned are different species from our study species. In the Pied flycatchers (Korneeva et al. 2006) experimental conditions were different: recordings were made from unanesthetized nestlings with implanted electrodes directly in the brain (field L), so likely with better SNR. The audiograms show a 10 dB SPL threshold after day 11, so the species may be considerably more sensitive than the zebra finch. The swamp sparrows in the Schroeder and Podos (2023) behavioral study were exposed for 4 days starting at 4-7 days post-hatch, so the study does not address embryonal hearing.

(d) zebra finch nestlings at 13d post-hatch, which show adult-like processing of songs in the auditory cortex (CNM) (Schroeder & Remage-Healey 2021, doi: 10.1002/dneu.22802).

(R3. A6d) This study does not conflict with our data. Even though sensitivity is lower at 10 days than in adults, cortical processing could still be ‘adult-like’.

(e) zebra finch juveniles, which are able to perceive and learn song syllables at 5-7kHz (fundamental frequency) with very similar acoustic properties to heat calls, and also produced during inspiration (Goller & Daley 2001, doi: 10.1098/rspb.2001.1805).

(R3. A6e) This result is not in conflict with our data. First, the onset of song learning occurs earliest at 20 DPH as discussed in the paper and our work demonstrates that click-evoked ABR thresholds are adult-like at 20 DPH. In the cited paper, the tutoring experiments were initiated at 35 DPH so the auditory system of the studied juveniles is mature.

Second, even though Goller & Daley 2001 do not report the source level of the specific syllable or the playback sound pressure levels, the source level of the inspiratory notes is comparable to other syllables, and thus around ~67 dB and ~34 dB louder than heat whistles.

NONE of these results - which contradict results and claims in the preprint - are mentioned.

Instead, the preprint focuses on very slow-developing species (parrots and owls), which take 2-4 times longer than songbirds to fledge (cited: Brittan-Powell et al 2004; Köppl & Nickel 2007; Kraemer et al 2017).

(R3. A6f) We have included papers in our discussion that reflect the known data (to our best knowledge) on the developmental neurophysiology and neuroanatomy of the auditory system and not proxies thereof.

(7) Results in figures are misreported in the text, and conclusions in the abstract and headers are not supported by the data:

For example:

(a) The data on Figure 1E shows that at 4 days old, 8 out of 13 nestlings (60%) responded to clicks, but the text says only 5/13 responded (L89).

(R3. A7a1) We apologize for this typo. Corrected.

When 60% (4dph) and 90% (6dph) of individuals responded, the correct term would be that "most animals", rather than "some animals" responded (L89).

(R3. A7a2) We have rephrased this sentence into “observable in most animals during” as suggested.

Saying that ABR to loud sound appeared "in the majority only after one week" (L93) is also incorrect, given the data.

(R3. A7a3) We have rephrased this sentence into: “Thus, sounds at loud, yet physiologically relevant SPLs do not evoke ABRs in the first days after hatching, but do so in all animals at 8 DPH.” (Lines 97-99).

It follows that the title of the paragraph is also erroneous.

(R3. A7a4) The paragraph title supports our conclusions and we will keep it.

(b) The hearing threshold is underestimated by 40dB at 6 and 8Kz on Fig 2C, not by "10-20dB" as reported in the text (L178).

(R3. A7b) We have changed the title of this section and moved the last sentence to the discussion to remove the focus on heat whistles. We added a paragraph in the discussion to specifically address the difference between ABR and behaviorally measured audiograms (Lines 413-422).

(B) Egg vibration experiment

(8) Using airborne sound to vibrate eggs is biologically irrelevant:

(R3. A8.1) We agree that parental contact could influence vibration transmission.

However, (1) prior studies assume airborne sound transmission, and (2) our experiment tests this assumption directly. We now clarify this scope (Lines 328-342 and Figure 4) and discuss contact-based transmission as a potential future direction.

The measurement of airborne sound levels to vibrate eggs misunderstands bone conduction hearing and is not biologically meaningful: zebra finch parents are in direct contact with the eggs when producing heat calls during incubation, not hovering in front of the nest. This misunderstanding affects all extrapolations from this study to findings in studies on prenatal communication.

(R3. A8.2) The definition of bone conduction is the response to sound that is not mediated by a functional middle ear, but through the skull. In the earlier study, the eggs were stimulated by sound from a headphone, so that is the reason for using the same stimulation here. See also joint response A3.5 above.

(C) Misrepresentation of current knowledge

(9) Values from published papers are misreported, which reverses the conclusions:

(R3. A9) We thank the reviewer for identifying inconsistencies and have:

- Corrected heat whistle frequency ranges consistently through our paper

- Added a comprehensive comparison figure gathering all available audiograms (Fig 5)

- Expanded discussion of high-frequency hearing.

These revisions do not alter our conclusions.

Most critical examples:

(a) Preprint: "Zebra finch most sensitive hearing range of 1-to-4 kHz (Amin et al., 2007; Okanoya and Dooling, 1987; Yeh et al., 2023)" (L173).

Actual values in the studies cited are:

1-to-7kHz, in Amin et al 2007 (threshold [=50dB with ABR] is the same at 7kHz and 1KHz).

1-to-6 kHz, in Okanoya and Dooling (the threshold [=30dB with behaviour] is actually lower at 6kHz than at 1KHz).

1-to-7kHz, in Yeh et al (threshold [=35-38dB with behaviour] is the same at 7kHz and 1KHz).

(R3. A9a.1) In this sentence presenting our results (“sensitive hearing range of 1-to-4 kHz”) we originally wrote that “these are consistent with the following papers (Amin et al., 2007; Okanoya and Dooling, 1987; Yeh et al., 2023)". This latter part was left out during the writing process. This explains the different numbers. We apologize for this mistake.

To avoid confusion, in our revision we have placed all ABR curves together into new Fig 5 and have included a new paragraph to discuss the differences (Lines 387-422).

Note that zebra finch nestlings' begging calls peaking at 6kHz (Elie & Theunissen 2015, doi: 10.1007/s10071-015-0933-6), would fall 2kHz above the parents' best hearing range if it were only up to 4kHz.

(R3. A9a.2) Of course that is possible. However this representation is incorrect because begging calls are harmonic sounds with a fundamental frequency around 500 Hz and formant at 6 kHz. Begging calls thus contain lots of energy at frequencies below 6 kHz, while the heat whistles do not. The peak frequency of heat whistles is also their lowest frequency component.

(b) The preprint incorrectly states throughout (e.g., L139, L163, L248) that heat-calls are 7-10kHz, when the actual value is 6-10kHz in the paper cited (Katsis et al, 2018).

(R3. A9b) The authors in Katsis et al. 2018 provided a range of 6-10 kHz estimated from the spectrogram without any further specification of methods. In another manuscript, we have quantified the heat whistle frequency (Anttonen et al Curr Biol https://doi.org/10.1016/j.cub.2025.08.054) to be 6.8 ± 0.6 kHz. We have changed this accordingly throughout our manuscript.

(c) Using the correct values from these studies, and heat-calls at 45 dB SPL (as measured by others (unpublished data), or as measured by the authors themselves, but which is not reported here (Anttonen et al 2025), the correct conclusion is that heat calls fall within the known zebra finch hearing range.

(R3. A9c) Please see our answer R3.A5a. We have included this argument more clearly in our discussion (Lines 271-355), illustrated by new Fig. 4.

(10) Published evidence towards high-frequency hearing, including in early development, is systematically omitted:

(a) Other studies showing birds use high frequencies above the known avian hearing range are ignored. This includes oilbirds (7-23kHz; Brinklov et al 2017; by 1 of the preprint authors, doi: 10.1098/rsos.170255) and hummingbirds (10-20kHz; Duque et al 2020, doi: 10.1126/sciadv.abb9393), and in a lesser extreme, zebra finches' inspiratory song syllables at 57kHz (Goller & Dalley, 2001).

(R3. A10a) We agree that some bird species produce or use acoustic signals extending into high frequencies. However, signal production is not evidence of perceptual sensitivity. Many animals, including birds and mammals, produce signals that contain harmonic or broadband components extending beyond their most sensitive hearing range without implying functional detection at those frequencies.

The cited examples (oilbirds, hummingbirds, inspiratory song syllables in zebra finches) concern signal production or ecological specializations in different species, not measured auditory sensitivity in zebra finches, and particularly not during early development. As such, they do not provide evidence that zebra finches—adults or embryos—can detect low-amplitude, narrowband signals in the 6–7 kHz range.

Our study explicitly addresses auditory sensitivity using physiological measurements, which is the relevant metric for evaluating detectability.

(b) The discussion of anatomical development (L228-241) completely omits the well-known fact that the avian basilar papilla develops from high to low frequencies (i.e., base to apex), which - as many have pointed out - is opposite to the low-to-high development of sensitivity (e.g., cited: Cohen & Fermin 1978; Caus Capdevila et al 2021).

(R3. A10b) We agree that the avian basilar papilla develops from base to apex (high to low frequency). We have now added a sentence in the Discussion to acknowledge this (Lines 406411).

Importantly, morphological development does not directly translate to functional sensitivity. Functional hearing depends critically on factors such as hair cell innervation, synaptic maturation, and central auditory processing, which are known to develop over time.

Our data show a low-to-high frequency progression in functional sensitivity, consistent with previous physiological studies. This apparent mismatch between anatomical gradients and functional onset has been noted in other systems and likely reflects the later maturation of neural encoding rather than hair cell differentiation per se. We now clarify this distinction in the revised manuscript (Lines 406-411).

(c) High frequency hearing in songbirds at hatching is several orders of magnitude better than in chickens and ducks at the same age, even though songbirds are altricial (e.g., at 4kHz, flycatcher: 47dB, chicken-duck: 90dB; at 5kHz, flycatcher: 65dB, chicken-duck: 115dB; Korneeva et al 2006, Saunders et al 1974). That is because Galliformes are low-frequency specialists, according to both anatomical and ecological evidence, with calls peaking at 0.8 to 1.2kHz rather than 2-6kHz in songbirds. It is incorrect to conclude that altricial embryos cannot perceive high frequencies because low-frequency specialist precocial birds do not (L250;261).

(R3. A10c) We agree that species differ in their auditory ecology and frequency specialization, and we do not claim that all altricial birds share identical developmental trajectories.

However, the cited comparisons involve different species, methodologies, and developmental timelines, which limits their direct comparability. In particular:

Developmental staging is not directly comparable across species using days post-hatch alone.

- Different methods (e.g., invasive recordings vs. ABR vs behavioural assays) yield systematically different thresholds.

- Ecological specialization (e.g., low-frequency vs. broadband species) influences adult audiograms and likely developmental trajectories.

We have revised the Discussion to explicitly acknowledge these limitations and to avoid overgeneralization across species. Importantly, our conclusions are based on within-species comparisons (adult vs. hatchling zebra finches) combined with measured signal levels of heat whistles. These constraints are sufficient to evaluate detectability without relying on cross-species extrapolation.

(11) Incorrect statements do not reflect findings from the references cited For example:

(a) "in altricial bird species hearing typically starts after hatching" (L12, in abstract), "with little to no functional hearing during embryonic stages (Woolley, 2017)." (L33).

There is no evidence, in any species, to support these statements. This is only a - commonly repeated - assumption, not actually based on any data. On the contrary, the extremely limited evidence to date shows the opposite, with zebra finch embryos showing ZENK activation in the auditory cortex in response to song playback (Rivera et al, 2018, not cited).

The book chapter cited (Woolley 2017) acknowledges this lack of evidence, and, in the context of song learning, provides as only references (prior to 2018), 2 studies showing that songbirds do not develop a normal song if the song tutor is removed before 10d post-hatch. That nestlings cannot memorise (to later reproduce) complex signals heard before d10 does not mean that they are deaf to any sound before day 10.

Studies showing hearing in young songbird nestlings (see point 6 above) also contradict these statements.

(R3. A11a) We agree that the precise onset of hearing in altricial embryos is not well established. We have therefore revised the wording in the Abstract and Introduction to avoid categorical statements and instead reflect the limited available evidence (Lines 13-16 and 33-37).

Our data provide direct physiological measurements showing extremely low sensitivity immediately after hatching, which constrains the likelihood of functional hearing in earlier embryonic stages.

Regarding the cited ZENK study, we note that immediate early gene expression indicates neural activation but does not provide a measure of auditory sensitivity or detection thresholds. As such, it cannot be directly compared to physiological or behavioral measures of hearing.

(b) "Zebra finch embryos supposedly are epigenetically guided to adapt to high temperatures by their parents high-frequency "heat calls" " (L36 and L135).

This is an extremely vague and meaningless description of these results, which cannot be assessed by readers, even though these results are presented as a major justification for the present study. Rather than giving an interpretation of what "supposedly" may occur, it would be appropriate to simply synthesize the empirical evidence provided in these papers. They showed that embryonic exposure to heat-calls, as opposed to control contact calls, alters a suite of physiological and behavioural traits in nestlings, including how growth and cellular physiology respond to high temperatures. This also leads to carry-over effects on song learning and reproductive fitness in adulthood.

(R3. A11b) We thank the reviewer for raising this point. In the revised manuscript, we have replaced the previous phrasing with a more precise and neutral summary of what these studies report, namely that embryonic exposure to heat-call playbacks has been associated with differences in physiological and behavioral traits.

Our study, however, addresses a distinct question—whether such acoustic signals are detectable by embryos given known constraints on signal amplitude and auditory sensitivity. The cited studies do not directly quantify auditory perception or the physical sound environment experienced by embryos. As a result, they do not provide a direct test of the sensory mechanism required for acoustic communication. A detailed evaluation of experimental design and interpretation in those studies is beyond the scope of the present manuscript, and we therefore limit our discussion to assessing the biophysical and physiological plausibility of the proposed mechanism.

(c) "The acoustic communication in precocial mallard ducks depends specifically on the lowfrequency auditory sensitivity of the embryo (Gottlieb, 1975)" (L253)

The study cited (Gottlieb, 1975) demonstrates exactly the opposite of this statement: it shows that duckling embryos, not only perceive high frequency sounds (relative to the species frequency range), but also NEED this exposure to display normal audition and behaviour post-hatch. Specifically, it shows that duckling embryos deprived of exposure to their own high-frequency calls (at 2 kHz), failed to identify maternal calls post-hatch because of their abnormal insensitivity to higher frequencies, which was later confirmed by directly testing their auditory perception of tones (Dimitrieva & Gottlieb, 1994).

(R3. A11c) We thank the reviewer for this clarification and have revised the relevant text. Our intention was to highlight that embryonic auditory experience can shape postnatal behavior, not to imply strict low-frequency limitation. Therefore we already included the actual frequency in the original sentence. We have removed the non-descriptive term “low-frequency” (Lines 330-332).

(12) Considering all of the mistakes and distortions highlighted above, it would be very premature to conclude, based on these results and statements, that altricial avian embryos are not sensitive to sound. This study provides no actual scientific ground to support this conclusion.

(R3. A12) We respectfully disagree with the reviewer’s conclusion.

Our study does not make a general claim that altricial embryos are incapable of perceiving sound. Rather, we evaluate a specific hypothesis: whether zebra finch embryos and hatchlings can detect sound and parental heat whistles.

Our conclusions are based on the convergence of:

(1) Measured low sound pressure levels of heat whistles,

(2) Established adult auditory thresholds (behavioral data),

(3) A large developmental decrease in auditory sensitivity demonstrated by our ABR measurements.

Even under conservative assumptions, these constraints place heat whistles at or below adult detection thresholds and far below expected sensitivity in hatchlings and embryos.

Thus, our conclusion is not based on absence of evidence, but on quantitative constraints that make detection unlikely under biologically realistic conditions.

Recommendations for the authors:

Joint recommendations:

In response to the joint recommendations, we have:

- Expanded methodological transparency (temperature, electrode setup, stimulus parameters),

- Added new data (Fig S3) and figures (Fig 4 and 5),

- Clarified ABR limitations and interpretation,

- Strengthened the separation between measured results and interpretation,

- Reframed conclusions to avoid overstatement.

These revisions leave the central two conclusions unchanged: 1) zebra finch hatchlings and embryos are functionally deaf, and 2) under biologically realistic conditions, heat whistles are unlikely to be detectable by zebra finch hatchlings or embryos.

(A) Reviewers 1 and 2:

Much of the reviewer discourse revolved around providing clarifications of methodology for measuring the ABR and caveats for interpretation. There was near consensus with reviewers 1 and 2 on issues related to the ABR, which should be addressed.

We appreciate the reviewers’ consensus that the main conclusions are supported, while requesting clarification of methodological details and interpretation of ABR measurements.

(1) Please address all of the issues raised by reviewers 1 and 2 above.

(A1) All points raised by Reviewers 1 and 2 have been addressed in detail in our point-by-point rebuttal below. In addition, we have revised the manuscript to improve clarity, added new figures (Fig. 4, 5), and substantially expanded the Discussion with eight new paragraphs to better contextualize our findings.

(2) Please also

- clarify all aspects of experimental details of the ABR that were missing, including temperature control (estimate body and ambient temperatures during ABR recordings,

- please address the possibility of hypothermia of hatchlings that could have reduced ABR responses,

- and potential local head cooling due to surgical exposure and its likely effect on highfrequency response depression).

(A2) In our revision, we have expanded the Methods section (L479-485) and added the following new data:

Body and ambient temperature/hypothermia

We have now included the body temperatures during ABR recordings in new table S6. These data show that:

(1) Body temperature was stable throughout recordings,

(2) Temperatures were within the physiological range,

(3) Conditions were consistent across all age groups.

Importantly, even the youngest hatchlings maintained stable temperatures and showed no indication of hypothermia. Therefore, differences in ABR responses cannot be attributed to temperature effects.

Potential cooling due to surgical exposure

This concern does not apply to our experiments. We used subdermal needle electrodes, which do not require surgical exposure. Therefore, no local cooling of the head occurred, and no tissue exposure could affect high-frequency sensitivity. We have added a clarifying sentence in the Methods section to explicitly state this (Line 501-503).

(B) Reviewer 3 also had additional requests for clarification that should also be addressed:

(3.1) Stimulus duration too long: The study used 25 ms tone bursts instead of the standard {less than or equal to} 5 ms "pips." Could this prevent reliable ABR detection, especially at high frequencies?

(A3.1) We agree that stimulus duration affects ABR characteristics and now clarify our rationale in the manuscript.

- The 25 ms tone bursts were deliberately chosen to ensure sufficient cycle representation at low frequencies (down to 250 Hz) and to avoid frequency splatter.

- Using a constant duration across frequencies ensures comparable stimulus energy.

Importantly:

- The 25 ms data yield audiogram shapes consistent with both click responses (Fig 2C) and published behavioral data (new Fig 5).

- To address potential high-frequency limitations, we included a dataset using 5 ms tone pips, which produced thresholds consistent with published ABR studies (new Fig 5).

Thus, both stimulus types support the same conclusion: a gradual maturation of hearing sensitivity from low to high frequencies. We have expanded the Discussion with three paragraphs to clarify these methodological trade-offs (Lines 387-422).

(3.2) Were 400 sweeps enough averaging? Might a signal appear at 1000 or more?

(A3.2) Signal-to-noise ratio improves with the square root of the number of averages. Increasing from 400 to 1000 sweeps would therefore reduce thresholds by at most ~4 dB. This magnitude is small relative to the >54 dB developmental differences observed, and the large gap between signal levels and detection thresholds. Thus, increasing sweep number would not alter the conclusions. We now clarify this explicitly in the Methods (L530).

(3.3) Was the repetition rate too high? How does the stimulus presentation affect the ABR? Might a signal have emerged with 1-4 per second?

(A3.3) We have clarified stimulus presentation rates in the revised manuscript:

- Clicks were presented at 25 Hz. Control measurements (now included as Supplementary Fig. S3) show no effect of this rate on ABR amplitude or threshold.

- Tone bursts and pips were presented at ~3 Hz, consistent with commonly used rates that avoid neural adaptation. We apologize for leaving this out in our original submission.

We now explicitly describe these parameters and their rationale in the Methods (Lines 543-552).

(3.4) If possible, provide an estimate of the effective bandwidth of the tone pips and compare it with the bandwidth of the parental heat-whistles.

(A3.4) We agree that stimulus bandwidth differs between tone pips and heat whistles, and that broader signals may stimulate multiple auditory filters. Shorter stimuli (e.g., 5 ms pips) have broader bandwidth and may stimulate multiple filters—particularly at low frequencies—potentially lowering thresholds, whereas longer stimuli (25 ms bursts) are more frequency-specific and may yield higher thresholds. At higher frequencies (including the heat whistle range), this effect is expected to be smaller.

However, quantitative correction is currently not possible due to a lack of species-specific data on auditory tuning curves in zebra finches. The only available avian data (budgerigar; Saunders et al., 1979) suggest auditory filter bandwidths (Q10 dB, i.e., the bandwidth 10 dB below the peak divided by peak frequency) of ~1.4 at low frequencies and ~1000 Hz at higher frequencies, but how multifilter stimulation affects thresholds is unknown and likely species- and frequency-dependent.

Given these uncertainties, direct comparison between tone stimuli and heat whistles requires strong assumptions. We therefore suggest that future studies should measure responses to natural heat whistles directly.

(3.5) Egg Vibration Experiment. Address the possibility that if a parent were physically lying on top of an egg and generated a heat call, parental body vibration could significantly communicate some perceptual vibrotactile signal to the egg. Reviewer 3 raised the possibility that the experiments in this paper tested the extent to which an auditory input can vibrate the egg - what if a vocalizing bird was on the egg?

(A3.5) We agree that embryos may receive multiple types of sensory input from parents, including direct mechanical cues.

However, our experiment specifically tests the hypothesis proposed in prior work: that airborne sound (heat whistles) induces egg vibrations sufficient for perception. Our findings show that airborne sound-induced vibrations are orders of magnitude below known vibrotactile sensitivity thresholds.

Regarding parental contact:

- Heat whistles are produced by an aerodynamic whistle mechanism, not tissue vibration (Anttonen et al., Curr Biol 2025), meaning most respiratory energy is radiated as sound rather than dissipated as heat/vibration in the body.

- A parent sitting on the egg would attenuate airborne sound transmission, not amplify it.

We now clarify in the Discussion that other cues (e.g., respiration, direct contact, temperature) may exist and need to be included in new experiments (Lines 349-352). Even so, these are distinct mechanisms and were not the hypothesis tested in prior playback studies.

In our paper, we will not add a detailed discussion of these prior papers as this is outside the scope of this paper. Instead, we added a paragraph what would be a constructive way forward (Lines 349-355). Hopefully somebody in the community will have the good fortune to secure research funding to continue this benchmarking work.

(4) Finally, all reviewers agreed that some more context on the ABR and its relationship to functional hearing could be provided, with less direct focus on the heat-call experiments.

(A4) We agree and in the revision discussion have compiled all published ABR and behavioral audiograms (Fig. 5) and added new paragraphs on functional hearing (Lines 261-269), and ABR vs behavioural audiograms (Lines 413-422).

Furthermore, to remove focus on the heat whistles, we have moved all heat-whistle-specific interpretation out of the Results into a single, focused Discussion section (Line 271-355).

The behavioral studies mentioned below show auditory responses (e.g., begging suppression). However, these behaviors are tested between ~5–10 days post-hatch (consistent with our findings), in different species, and do not provide quantitative sensitivity thresholds, nor do they address detectability of low-amplitude, high-frequency signals like heat whistles.

In our revision we added a new discussion paragraph including these behavioral studies (Lines 357-367).

For example, there are ample cases in the literature of altricial birds exhibiting behavioral evidence of auditory sensitivity by reducing begging calls in response to parental alarm calls:

Platzen & Magrath (2004) - Playback of parental alarm calls nearly abolished nestling non-begging calls and reduced begging in scrubwrens. Proc. R. Soc. B 271:1271-1276.

Different species: scrubwrens. Playback age: 5-, 8- and 11-DPH nestlings.

Magrath, Haff, Horn & Leonard (2006) - Review and experiments on the developmental shift to silence/freeze after aerial alarm calls as chicks become fledglings; documents nestling quieting to alarms. Proc. R. Soc. B 273:2335-2341.

Different species: scrubwrens. Playback age: 7-9 DPH nestlings, and 2- 4 days after fledging.

Magrath, Pitcher & Dalziell (2007) - Nestlings respond to the sound of a predator's footsteps and parental food/alarm calls; includes begging suppression following predator sounds. Anim. Behav. 74:1117-1129.

Different species: scrubwrens. Playback age: 8 DPH nestlings.

Haff & Magrath (2012) - Nestlings suppress calling after heterospecific alarm calls (when acoustically similar to conspecific alarms), indicating generalized auditory danger recognition. Anim. Behav. 84:e.g., 495-505 (article).

Different species: scrubwrens. Playback age: 5-6 and 10-11DPH nestlings. They show that 10-11 days old suppress calling while 5-6 days old do not.

Barati & McDonald (2017) - Noisy miner nestlings suppress begging after conspecific alarm calls and some heterospecific cues; stronger/longer suppression for terrestrial-predator alarms. Sci. Rep. 7:9563.

Different species: Noisy miner (Manorina melanocephala ). Playback age: 14 DPH. Nestlings started to vocalise at 5 DPH.

Suzuki (2011) - In Paridae, parental alarm calls encode predator type; prior work (cited within) shows young of altricial species suppress vocalizations to alarms. Curr. Biol. 21:15-20.

Different species: great tits. Playback age: 17 DPH.

Can you please contextualize the present results about the timing of auditory development with the above body of work with respect to the timing of alarm call-induced begging call suppression?

In our revision we have added a new paragraph in the discussion on these papers (Lines 357-367), and highlight the need for comparative work on hearing development in different species (Lines 352-355 and Lines 405-411).

(5) Strictly speaking, a flat ABR does not equal deafness - at the extreme, an average of 10,000 trials may pull out a minuscule signal. Thus, the more rigorous path would be, in the results section, to ensure that statements summarize the data as they are, representing an absence of a neural signal.

Save the interpretation of what this may mean for the discussion, and provide alongside this interpretation the necessary caveats related to temperature, rendition rate, averaging, etc.

Clarify the conditions where a flat ABR demonstrates or fails to demonstrate immature deafness.

Expand clarification for how the known 20-40 dB difference between ABR and behavioral thresholds can exist if a flat ABR can be interpreted as deafness.

Consider refraining from concluding deafness from a flat ABR. Discuss that behavioral, single-unit, or alternative physiological assays might detect responses below the ABR threshold. If such cases exist, cite.

(A5) We thought about this considerably before starting our measurements. What constitutes the absence of a signal? Even with intracellular recordings of all but one of the auditory neurons, the last one could still contain a signal and theoretically transmit information to the nervous system. We agree with the reviewers that absence of an ABR response should not be equated with absolute deafness. We have revised the manuscript accordingly and removed all statements implying “deafness” from the Results. The Results now strictly report presence or absence of detectable ABR responses.

However, in both clinical and comparative contexts, absence of ABR responses at high SPLs (e.g., 90–95 dB) is widely interpreted as functionally non-responsive hearing. The developmental shift we observe (>54 dB) is far larger than typical ABR–behavioral offsets (20–40 dB). In the Discussion, we have added a new paragraph arguing that we think that the term functional deafness is reasonable here (Lines 261-269).

Version 2

Abstract

Acoustic communication relies critically on the receiver’s ability to hear. In precocial birds, hearing can already be functional during embryonic stages in the egg, whereas in altricial species the limited available data indicate a gradual postnatal development with poor sensitivity in early hatchlings, implying even lower sensitivity in embryos. Recent research suggests that zebra fnnch embryos, despite being altricial, have functional hearing already in the egg to engage in parent-embryo acoustic communication and anthropogenic noise detection. However, their auditory sensitivity during early development remains unknown. Here, we measure auditory brainstem responses (ABR) over early postnatal development and show that zebra fnnch hatchlings have >54 dB lower threshold than adults to loud, broadband clicks. ABR responses emerge between 4-8 days after hatching. Auditory sensitivity develops progressively and reaches adult levels by day 20. Additionally, egg vibrations induced by sound remain far below detection thresholds of vibrotactile senses. Together these data contradict the notion of acoustic perception in embryos and early parent-to-embryo communication in zebra fnnches. The striking timing coincidence between maturation of the peripheral auditory system and the onset of song learning suggests that hearing functionality may gate the onset of vocal learning in zebra fnnches.

Introduction

The ability to detect sound is critical for vocal communication, species recognition, predator detection, and survival across the vertebrates (Nordeen and Nordeen, 1992; Heaton et al., 1999; Zevin et al., 2004; Ladich and Winkler, 2017). In birds, precocial species that are able to move around after hatching show some hearing function already inside the egg (Gottlieb, 1965; Saunders et al., 1973; Konishi, 1973). This enables embryo-to-embryo (Vince, 1966; Woolf et al., 1976; Schwagmeyer et al.,1991; Noguera and Velando, 2019) and parent-to-embryo communication that is critical to imprinting on species-specifnc sounds (Gottlieb, 1965; Grier et al., 1967; Hess, 1972). In contrast, altricial birds hatch considerably earlier in their development and rely on parental care to survive (Orkney et al. 2025). The sparse data available show indicate a gradual postnatal development of hearing sensitivity, with poor sensitivity in early hatchlings (Brittan-Powell and Dooling, 2004; Korneeva et al., 2006; Aleksandrov and Dmitrieva, 1992; Kraemer et al., 2017), implying even lower sensitivity in the less developed prenatal embryos.

Remarkably, recent research directly challenges this concept and suggests that altricial songbirds can detect soft sounds while still in the egg (from here onwards referred to as embryos). In zebra fnnches, embryonic exposure to high-frequency “heat-call” (from here onwards referred to as heat whistles (Anttonen et al., 2025) playbacks has been associated with differences in physiological and behavioral traits (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023) and prehatching anthropogenic noise exposures reduced embryonic survival and reproductive success in adulthood (Meillère et al., 2024). Furthermore, fairy wren embryos supposedly learn elements of the female incubation call already at 10-12 days after fertilization (Colombelli-Négrel et al., 2012, 2014, 2016; Kleindorfer et al., 2024). The above interpretations critically depend on the ability of these embryos to detect soft level sounds in the egg and after hatching. However, it is currently unknown whether these species can detect sounds this early in their development.

The zebra fnnch is a major animal model system for studying the neural circuitry underlying hearing-guided behaviors including vocal imitation learning. However, we know surprisingly little about the development and the function of the auditory organ in this species. In singing males, the onset of sensory and sensorimotor learning starts around 25 days-post-hatch (DPH) and continues to adulthood at ∼90 DPH (Immelmann, 1967; Böhner, 1990; Roper and Zann, 2006; Braaten, 2010; Gobes et al., 2019; Leitão and Gahr, 2024). Aligned with the onset of this behavior, the overall hearing sensitivity is comparable to adult levels at 20 DPH (Amin et al., 2007) with a frequency range from 0.25 to 8 kHz and highest sensitivity around 1-6 kHz (Behavior 1-6 kHz: Okanoya and Dooling, 1987; Yeh et al., 2023; ABR 1-4 kHz, Amin et al., 2007; Noirot et al., 2011; Zevin et al 2004). However, at 10 DPH - the earliest age measured – hearing sensitivity is reduced by 25 dB compared to 20 DPH and is thus ∼20-fold lower (Amin et al., 2007). Because hearing sensitivity in zebra fnnches increases progressively in late postnatal development - as in other bird species (Rebillard and Rubel, 1981; Katayama, 1985; Jones et al., 2006; Brittan-Powell and Dooling, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017) their ability to detect sounds as embryos and as early hatchlings may be severely compromised. Furthermore, recent work shows that heat whistles are extremely soft sounds in vivo with a source level of only ∼14 dB re. 20 µPa at 1 meter (sound pressure level of ∼34 dB re. 20 µPa at 10 cm; Anttonen et al., 2025). Therefore, zebra fnnch embryos would require highly sensitive hearing function to detect such soft sounds. However, we lack conclusive evidence on the hearing capabilities of juvenile zebra fnnches.

Here we test the hypothesis that zebra fnches can hear during early postnatal development by measuring auditory brainstem responses (ABRs) that allow hearing threshold estimation when behavioral training is not possible (He et al., 2008). Furthermore, we test whether intense sounds can induce egg vibrations of sufficient magnitude for vibrotactile detection. We show that the earliest ABRs evoked by intense (95 dB), broadband sounds could not be detected on 2 DPH, but earliest at 4-8 DPH. The threshold at 4DPH is at least >54 dB higher compared to adult levels. Responses to frequency specifnc sounds develop later in a progressive, low-to-high frequency order. ABR sensitivity to clicks matures to adult-like levels at 10 DPH but the response amplitude continues to increase beyond 20 DPH. Airborne sounds induce low amplitude vibrations in eggs that are far below typical avian and mammalian vibrotactile detection thresholds. Together, our fnndings show that zebra fnnch embryos and early hatchlings are functionally deaf and provide evidence against the hypothesis that they can detect sound stimuli via hearing or via vibrotactile senses. Our data furthermore show that auditory nerve responses continue to mature past 20 DPH, suggesting that the maturation of hearing limits the onset of vocal learning in zebra fnnches.

Results

Responses to loud clicks appear during the fnrst postnatal week

To assess the development of zebra fnnch hearing sensitivity in early postnatal development we quantifned the responses of auditory pathway neurons to sound stimuli by recording ABRs (Fig. 1A). First, to evaluate overall hearing sensitivity we presented loud yet physiologically relevant (95 dB SPL) clicks that stimulate the auditory system across a wide frequency spectrum (Fig. 1B). Clicks activate most auditory neurons nearly simultaneously and result in prominent ABRs, thus providing maximal sensitivity for threshold detection.

Click-evoked ABRs appear and gradually mature after hatching in zebra finches.

(A) ABR recording setup. (B) Used sound stimuli. Red, phase inverted signal. (C) Representative, 95 dB SPL click-evoked ABR ante-(black) and post-mortem (grey). (D) Representative ABRs to 95 dB SPL clicks at tested DPH timepoints. The first negative peak (I) and the following trough are designated as wave I. Noise RMS is determined from the red area and signal RMS from the blue area. (E) Percentage of animals that exhibited a detectable (S/N > 2) click-evoked ABR. (F) Detection thresholds were measured with a set of clicks with decreasing 5 dB SPL steps. Arrow indicates the determined threshold for this 8 DPH example. (G) Average click thresholds at different DPH ages. (H) Click-evoked ABR wave I amplitude and (I) latency over SPL. Age groups are color-coded in D-I. Shaded areas in H-I = s.e.m.

Clicks induced robust neural potentials in adult zebra fnnches (Fig. 1D-E). These potentials were not recording artefacts as they occurred only when the animal was alive (N = 3, Fig. 1C). None of the tested animals displayed click-evoked ABRs at 2 DPH (both with detection at signal-to-noise ratio of 2 (S/N > 2) and with detailed visual inspection; n=0/15; Fig. 1D,E). However, click-evoked ABRs became observable in most animals during the fnrst postnatal week (S/N > 2; 4 DPH, n = 8/13; 6 DPH, n = 9/10; Fig. 1D,E). Subsequently, all tested animals consistently demonstrated click-evoked ABRs from the second postnatal week onwards (S/N > 2; 8 DPH, n = 11; 10 DPH, n = 11; 20 DPH, n = 7; 25 DPH, n = 8; Adult, n = 8; Fig. 1D,E). Thus, sounds at loud, yet physiologically relevant SPLs do not evoke ABRs in the fnrst days after hatching, but do so in all animals at 8 DPH.

Click-evoked ABR thresholds are adult-like at 10 DPH

To estimate hearing sensitivity thresholds over postnatal age, we recorded series of click-evoked ABRs at 5 dB downward steps from 95 dB SPL (Fig. 1F). Hearing thresholds are strongly affected by age of the animal (F(6) = 25.22, p = 4.93*10-14; Fig. 1G, Table S1). Click-evoked ABR thresholds are signifncantly higher than adult-like levels (40.6 ± 4.0 dB SPL) in animals younger than 10 DPH but not in older animals (4 DPH vs. adult, t = 8.16, p = 3.27*10-10; 6 DPH vs. adult, t = 5.92, p = 1.39*10-6; DPH 8 vs. adult, t = 3.37, p = 0.008; 10 DPH vs. adult, t = 1.23, p = 1; 20 DPH vs. adult, t = -0.13, p = 1; 25 DPH vs. adult, t = -1.29, p = 1; p-values are Bonferroni-corrected; Fig. 1G, Table S1). Click thresholds are above 95 dB on 2 DPH and as high as 80.6 ± 1.6 dB SPL on 4 DPH. The threshold of a 2 DPH animal is at least 54 dB higher compared to an adult. Thus, zebra fnnch hearing sensitivity increases rapidly but gradually during the fnrst two postnatal weeks.

ABR wave I gradually reaches maturity 25 days after hatching

The click-evoked ABR consists of multiple wave peaks and troughs that reflect the subsequent activations of various neural populations of the auditory pathway. The fnrst wave (wave I in Fig. 1D) is considered to correspond to the activity of the primary auditory neurons (Brittan-Powell and Dooling 2002). To study the postnatal maturation of this synchronized neural output, we measured changes in both the amplitude and the latency of wave I over postnatal development.

In adults, the amplitude of wave I increases signifncantly with click SPL (Fig. 1H). Age signifncantly affects both wave I amplitude directly and the relation between SPL and wave I amplitude (Age-SPL interaction compared to only additive effects of age and SPL: 𝜒2(6)=164.23, p=7.52*10-33, additive effects of age and SPL compared to only SPL: 𝜒2(6)=79.43, p=4.69*10-15, SPL only compared to null model: 𝜒2(1)=438.12, p=2.78*10-97, Fig. 1H, Table S2). The steepness of the amplitude-SPL slope gradually increases with age and is signifncantly different between adults and younger animals, but not between adult and those aged 25 DPH (4 DPH vs. adult, t = -5.40, p = 1.50*10-6; 6 DPH vs. adult, t = -5.65, p = 1.8910-6; DPH 8 vs. adult, t = -6.30, p = 3.79*10-7; 10 DPH vs. adult, t = -5.53, p = 118 6.53*10-6; 20 DPH vs. adult, t = -3.00, p = 0.03; 25 DPH vs. adult, t = -0.69, p = 1; p-values are Bonferroni-corrected; Fig. 1H, Table S2).

The latency of wave I decreases with increasing click SPL (Fig. 1I). Furthermore, age signifncantly affects the relation between dB SPL and wave I latency (Age-SPL interaction compared to only additive effects of age and SPL: 𝜒2(6) = 116, p = 1.31*10-22; additive effects of age and SPL compared to only SPL, 𝜒2(6) = 109, p = 2.76*10-21; SPL only compared to null model: 𝜒2(1) = 754, p = 4.81*10-166; Fig. 1I, Table S3). The observed latencies are signifncantly longer during early postnatal development (4-to-10 DPH) when compared to latencies observed in adults. No signifncant differences are observed between the latencies of adults and animals older than 10 DPH (4 DPH vs. adult, t = 12.56, p = 3.41*10-19; 6 DPH vs. adult, t = 10.06, p = 1.49*10-13; 8 DPH vs. adult, t = 7.24, p = 1.16*10-8; 10 128 DPH vs. adult, t = 6.18, p = 5.95*10-7; 20 DPH vs. adult, t = 2.24, p = 0.18; 25 DPH vs. adult, t = 1.67, 129 p = 0.61; p-values are Bonferroni-corrected; Fig. 1I, Table S3).

Taken together, the synchronized output of the auditory nerve increases in sensitivity and reduces in latency postnatally progressively from 4 DPH until reaching maturity at 20-25 DPH. Such gradual increase is consistent with postnatal hearing development in budgerigars, mice, and humans (Brittan-Powell and Dooling, 2002, 2004; Hecox and Galambos, 1974; Henry and Haythorn, 1978).

Hearing sensitivity develops progressively from low to high frequencies

To study the postnatal development of frequency-specifnc hearing sensitivity we tested whether loud (95 dB SPL) tone bursts centered at frequencies between 0.25-to-8 kHz elicit detectable ABRs. Tone burst-evoked ABRs were fnrst detectable on 8 DPH at 1 kHz (Fig. 2A and Fig. S1A). As observed with click stimuli, the amplitude of tone burst-induced ABRs increases with age (Fig. 2A, Fig. S1B, Table S4; Age as a predictor of amplitude compared to null model: 𝜒2(3) = 6.11, p = 0.11). The latency of the response decreases signifncantly with age (Fig. 2a, Fig. S1c, Table S4; Age as predictor of latency compared to null model: 𝜒2(3) = 33.72, p = 2.27*10-7). Latencies are signifncantly longer at 8-to-10 DPH than at adulthood (1 kHz: 8 DPH, 6.60 ± 0.77 ms; 10 DPH, 5.01 ± 0.40 ms) and are adult-like at 20-to-25 DPH (1 kHz: 20-to-25 DPH, 3.34 ± 0.18 ms; adult, 3.67± 0.30 ms; 8 DPH vs. adult, t = 5.95, p = 1.87*10-6; 10 DPH vs. adult, t = 4.52, p = 0.0004; 20-to-25 DPH vs. adult, t = 1.16, p = 0.81; p-values are Bonferroni-corrected; Fig. S1C and Table S4). During the following weeks of postnatal development, tone burst-induced ABRs became detectable at 0.5-to-4 kHz (Fig. S1A). No responses were detected at higher frequencies (Fig. 2B,C; Fig. S1A). Taken together, the maturation of hearing sensitivity in zebra fnnches progresses in a low-to-high frequency order as is also observed for example in mice (Ehret, 1976), cats (Ehret and Romand, 1981), and in some other bird species (Brittan-Powell and Dooling, 2002, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017).

Sound frequency sensitivity matures over postnatal development.

(A) Representative ABRs to 1 kHz and (B) 8 kHz tone burst stimuli over postnatal development. (C) Zebra finch ABR thresholds obtained with tone bursts (solid lines) and with tone pips (dashed line).) Shaded areas = s.e.m.

Tone pip-evoked ABR thresholds

Tone bursts evoked ABRs only in about half of the adults (Fig. S1A) and the detection thresholds of > 60 dB SPL are high compared to earlier ABR studies (Amin et al., 2007; Noirot et al., 2011; Zevin et al 2004). This indicates that while being more frequency specifnc, tone bursts failed to recruit auditory neurons to fnre concurrently enough to generate detectable ABRs. Therefore, we next used shorter 5 ms-long tone pips (Fig. 1B) that have a broader spectral distribution of acoustic energy compared to tone bursts, but produce a more focused temporal response (Lauridsen et al., 2021).

Indeed, the SPL matched tone pips produced much more prominent ABRs (Fig. 2C and Fig. S2A) that were readily detected in all tested adults around their most sensitive hearing range of 1-to-4 kHz (Fig. S2B). Detection thresholds for tone pips were signifncantly lower than those of tone bursts with ∼20 dB SPL (Additive effects of stimulus type and frequency as predictors of threshold compared to stimulus type alone χ2(1) = 6.84, p = 0.009; stimulus type as a predictor of threshold compared to null model: χ2(1) = 10.27, p = 0.001, Table S5).

Sound-induced egg vibrations

These data suggest it unlikely that embryonic zebra fnnches can detect soft, high frequency heat whistles as airborne sounds via their auditory system (See Discussion). Therefore, we next tested if airborne sounds can induce zebra fnnch eggs to move at sufficient velocities to be detected as vibrations.

We played sound frequency sweeps to eggs while simultaneously recording the sound-induced vibrations of the egg with a laser vibrometer (Fig. 3). The vibration velocity transfer function shows that eggs vibrate at velocities around -85 to -55 dB re 1 mm s-1 when exposed to a loud 94 dB SPL frequency sweep between 0.25-to-10 kHz (Fig. 3C). Within the frequency ranges associated with zebra fnnch song (1-4 kHz, Elie and Theunissen 2016) and heat whistles (6.8 ± 0.6 kHz (Anttonen et al, 2025), sound exposures at 94 dB SPL induce average vibration velocities of -72.4 ± 0.8 and -70.7 ± 1.2 dB re 1 mm s-1(N=5), respectively (Fig. 3B). We calculated the corresponding vibration velocity levels induced by SPLs used during experimental heat whistle playbacks to zebra fnnch eggs (62-to-67 dB SPL)(Katsis et al., 2018) to be between -99.4 and -96.2 dB re 1 mm s-1, respectively (Fig. 4).

Intense sound stimulation induces zebra finch eggs to vibrate at very low velocities.

A Setup used to measure sound induced vibrations of eggs. A 94 dB, 0.25-to-10 kHz frequency sweep was played at the eggs to determine the vibration transfer function. B Egg vibration velocities at frequencies associated with song (left) and with heat whistles (right). Error bars = s.d. C Example vibration velocity transfer function of a zebra finch egg (mean ± S.D. (grey area) N=5) Green and red areas display the frequency ranges of song (Elie and Theunissen 2016) and heat whistles (Anttonen et al, 2025), respectively, that were used to calculate mean velocity in panel B.

Heat whistles are below hearing and vibration threshold levels in hatchlings and adults.

A In vivo heat whistles at 10 cm distance (Anttonen et al., 2025) (orange) are just on or below the adult hearing threshold based on behavioral audiograms (blue line; Yeh et al. 2023; green line, Okanoya and Dooling, 1987). Increasing the most sensitive adult audiogram by 54 dB as a conservative estimate of reduction based in clicks (Fig 1G), shows that both in vivo heat whistles and previously playback levels (Katsis et al., 2018) lay far below the estimated thresholds for a 2DPH old hatchling, even without an expected further increased high frequency threshold (horizontal arrow). Thresholds for embyos still in the egg are likely be shifted further up, and sound dampening by parents incubating the egg will reduce the heat whistle levels further down, making the gap even larger. B Egg vibration velocities caused by 67 dB song, 65 dB heat whistle playbacks (Katsis et al., 2018), and in vivo heat whistles (Anttonen et al., 2025) are far below with thresholds for human (Håkansson et al., 1985) and mouse bone conduction (BC) hearing (Chhan et al., 2017) and for human vibrotactile threshold at 0.2 kHz (Mountcastle et al., 1972).

Discussion

We show that hearing functionality gradually increases over postnatal development in zebra fnnches. An evoked response (ABR) to loud click stimuli, reflecting the synchronized output of auditory nerve fnbers, is fnrst detectable at 4-8 days post hatch (DPH) after which hearing sensitivity gradually increases until reaching adult levels at 20-25 DPH. The response to longer tone burst also gradually increases consistent with hearing development in altricial and precocial bird species (Saunders et al., 1973; Konishi, 1973; Rebillard and Rubel, 1981; Katayama, 1985; Jones et al., 2006; Brittan-Powell and Dooling, 2004; Köppl and Nickel, 2007; Kraemer et al., 2017; Aleksandrov and Dmitrieva, 1992; Lippe, 1994; Amin et al., 2007) - as well as in mammals (Henry and Haythorn, 1978; Ehret and Romand, 1981). Therefore, the functional maturation of hearing sensitivity over postnatal development in zebra fnnches follows a gradual, progressive pattern consistent with other birds and mammals.

The progressively increasing hearing sensitivity over postnatal development (Fig. 1) implies that zebra fnnch embryos have a further reduced hearing sensitivity compared to early hatchlings. Even in precocial bird species that show auditory function already in the egg, auditory thresholds remain very high until a few days before hatching (Saunders et al., 1973; Rebillard and Rubel, 1981; Lippe, 1994; Saunders et al., 1974). In precocial chickens, evoked responses to very loud sounds (>100 dB SPL) have been obtained around embryonic day 12 (chickens hatch at day 21) (Saunders et al., 1973), but ambient sound levels evoke responses around four days later (Jones et al., 2006). Importantly, these measurements involved experimentally opening the egg, clearing the auditory meatus from fluids, and even drainage of fluids from the middle ear (Rebillard and Rubel, 1981) and thus may overestimate the natural hearing sensitivity. Although ABR provides easily accessible measurements of hearing sensitivity, the technique does not inform on the anatomical and neural processes underlying the gradual development of hearing sensitivity. In birds, these processes have been studied in most detail in chickens, where fluid unloading and maturation of middle ear structures causes signifncant improvements in hearing sensitivity during development (Saunders et al., 1973; Cohen et al., 1992). The onset of evoked responses to very loud sounds furthermore coincides with synapse formation on hair cells of the inner ear (Saunders et al., 1973; Cohen and Fermin, 1978; Caus Capdevila et al., 2021). Furthermore, the response characteristics of the auditory nerve fnbers and the latencies of the ABR show postnatal changes (Katayama,1985; Jones et al., 2006; Manley et al., 1991). The periphery of the chicken auditory system also continues to anatomically mature for several weeks after hatching (Cohen et al., 1992; Ryals et al., 1984; Cotanche and Corwin, 1991). Obviously, precocial species hatch much later in their development than altricial species and when these hallmark events occur in zebra fnnch development remains unknown. Further work is required to understand the physiological basis for postnatal development of hearing sensitivity in the zebra fnnch and other songbird species.

Our data indicate that zebra fnnch embryos and early hatchlings are functionally deaf within the tested stimulus range. ABR measures synchronized population activity and therefore provides a conservative estimate of sensitivity (Pickles, 2012). In clinical practice, the absence of ABR responses at high intensities is used to diagnose severe or profound hearing impairment in human infants (Stapells, 2000; WHO, 2021), with functional deafness typically defnned at thresholds ≥81 dB. By this standard, the lack of detectable responses at the highest stimulus levels tested (95 dB SPL) reflects a profound reduction in functional sensitivity. While sparse activity below ABR detection thresholds cannot be excluded, such activity is unlikely to support meaningful perception or behavior. We therefore use “functional deafness” to denote the absence of detectable, population-level neural responses under physiologically relevant conditions, rather than absolute absence of any neural activity.

When in a hot environment, zebra fnnch parents emit heat whistles that are soft, high frequency sounds (∼7 kHz, Anttonen et al., 2025; 6-10 kHz, Katsis et al., 2018) compared to zebra fnnch song, that are produced during the high respiratory flow conditions consistent with thermal panting (Anttonen et al., 2025; McDiarmid et al., 2018; Pessato et al., 2020). Embryonic exposure to heat whistle playbacks has been associated with differences in physiological and behavioral traits (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al., 2021; Udino and Mariette, 2022; Katsis et al., 2023). Our data provide no support for the hypothesis that zebra fnnch embryos or early hatchlings can detect heat whistles via either auditory or vibrotactile pathways (Fig 4). This conclusion does not rely on a single dataset or method, but on the convergence of three independent constraints: (i) heat whistle sound level, (ii) adult auditory sensitivity, and (iii) developmental immaturity of the auditory system (Fig 4a). First, heat whistles are low-amplitude sounds. In vivo measurements show sound levels of ∼33 dB SPL at 10 cm and ∼14 dB at 1 m, with a conservative upper bound of ∼40 dB SPL even at very close range (<5 cm) (Anttonen et al., 2025). Second, behavioral audiograms indicate that adult zebra fnnches have thresholds of ∼30 dB SPL around 7 kHz, increasing steeply at higher frequencies (Okanoya and Dooling, 1987; Yeh et al., 2023) (Figs 4, 5). Thus, even under optimal conditions and at close range (10 cm), heat whistles lie at or below the detection threshold of adults (Fig 4A). Third, auditory sensitivity in early development is substantially reduced: our ABR data show a ≥54 dB loss in sensitivity in hatchlings relative to adults for click stimuli. In addition, the progressive frequency-dependent development from low to high suggests that the high frequency (6–8 kHz) thresholds can be expected to be even higher (left frequency shift in Fig 4A). Together, these constraints defnne a narrow and unfavorable detection window—a low-amplitude, high-frequency sound positioned at the edge of adult sensitivity, combined with a large developmental reduction in sensitivity—making detection by embryos or hatchlings highly unlikely. Importantly, this conclusion is robust to uncertainties in ABR–behavioral offsets; even under the most conservative assumptions, the sound remains at or below adult detection limits and far below those expected in early development.

Consistent with this, comparative data show that high-frequency hearing matures late across vertebrates, and even precocial species with embryonic auditory function exhibit very poor sensitivity above 6 kHz prior to hatching (Saunders et al., 1973; Katayama, 1985). Established parent-embryo communication in precocial mallard ducks depends specifncally on the auditory sensitivity of the embryo to detect maternal calls with frequency components below 3 kHz (Gottlieb, 1975). We further rule out a vibrotactile or bone-conduction pathway (Fig 4B): sound-induced egg vibrations by heat whistle playbacks (65 db SPL (Katsis et al., 2018)) are 10–40 dB below vertebrate bone-conduction thresholds and ∼90 dB below vibrotactile detection thresholds (Håkansson et al., 1985 (human bone conduction); Mountcastle et al., 1972 (human and monkey vibrotactile); Hörster, 1990 (pigeon vibrotactile)). In vivo heat whistles are at least 27 dB weaker than the used playback levels (Anttonen et al., 2025). Even assuming lossless transmission to the embryo, these vibrations remain orders of magnitude below known detection limits. Moreover, parental incubation would attenuate airborne sound while generating substantially stronger mechanical cues through respiration and contact, which were not the signals proposed to mediate these effects. Taken together, both auditory and mechanical pathways fall well below physiological detection limits, and are not plausible mechanism by which heat whistles could be perceived by embryos under natural conditions.

Taken together, our results provide evidence against the hypothesis that zebra fnnch embryos can hear sounds, and we consider it thus unlikely that adult zebra fnnches can communicate with their embryos via air-borne sound. How the heat whistle and anthropogenic noise playbacks have affected the embryos in the earlier studies (Mariette and Buchanan, 2016; Mariette et al., 2018; Katsis et al., 2018; Mariette and Buchanan, 2019; Pessato et al., 2020; Udino et al., 2021; Katsis et al.,2021; Udino and Mariette, 2022; Katsis et al., 2023; Meillère et al., 2024) remains to be further tested. Resolving this issue will require experiments that integrate realistic in vivo sound levels with simultaneous measurement of auditory, vibrotactile, and thermal cues. Playback studies should be calibrated to natural sound amplitudes (Anttonen et al., 2025), and complemented by approaches with higher sensitivity than ABR, such as single-unit recordings or optical imaging. A comparative framework spanning altricial and precocial species, combined with developmental staging based on anatomical and neural maturation of the auditory system and novel behavioral essays, will be essential to defnne the limits of sensory function in the avian embryo.

Behavioral studies show that altricial songbird nestlings suppress begging in response to parental alarm calls and predator cues several days after hatching, typically emerging between ∼5–10 days post-hatch and strengthening thereafter (Platzen and Magrath, 2004; Magrath et al., 2006; Magrath et al., 2007; Haff and Magrath, 2012; Barati and McDonald, 2017; Suzuki, 2011). This timeline closely matches the onset and maturation of ABR responses observed here in zebra fnnches. Importantly, these behaviors likely rely on detection of salient, often low-frequency or broadband cues, and do not imply mature auditory sensitivity. Korneeva et al. (2006) and Aleksandrov and Dmitrieva (1992) report ABR responses in 3DPH flycatcher hatchlings using invasive recordings in unanesthetized birds, which likely yield lower thresholds. Even so, sensitivity improves rapidly postnatally, with ∼40 dB threshold decreases and an upper frequency shift from ∼4 to ∼7 kHz within days. Our physiological data therefore align with the literature, indicating that ecologically relevant auditory function emerges only after an initial postnatal period of markedly reduced sensitivity.

Our adult pip audiograms are broadly consistent with values reported across multiple laboratories (Fig. 5), despite substantial methodological variation (Amin et al., 2007; Noirot et al., 2011; Zevin et al 2004). Differences in electrode confnguration (subdermal vs. in-brain), stimulus duration and envelope, repetition rate, and threshold criteria are all known to influence ABR thresholds and amplitudes (Stapells, 2000; Brittan-Powell et al., 2002), yet the results from four laboratories with different setups, conditions and bird populations are highly comparable. The remaining spread across studies likely reflects both technical and biological sources, including colony-specifnc variation and the limited representation of wild-derived zebra fnnches. Standardized protocols across populations will be required to disentangle methodological from biological variability in avian ABR datasets.

ABR and behavior-based audiograms in adult zebra finches are consistent in sensitivity and shape.

The ABR based audiograms from four different laboratories including our pip-data are consistent in overall sensitivity and shape. The distinct low threshold at 6 kHz in the data by Amin et al. (2007) may reflect real biological variation between laboratory populations. The behavioral-based audiograms (Okanoya and Dooling, 1987; Yeh et al., 2023) are also consistent in shape and thresholds. The systematically lower threshold of the behavior-based audiograms varies from about 20 dB (downward arrows) in the 0.25 – 4 kHz range to 20-40 dB at 6 kHz.

Tone-burst thresholds in our dataset are higher than click-evoked thresholds, consistent with reduced neural synchrony for longer, frequency-specifnc stimuli, particularly at higher frequencies. The sensitivity in responses between tone pips and burst may be caused by different rise times. Our 5 ms tone pip stimuli had rise-fall times of 1 ms based on procedures in earlier studies (e.g. Brittan-Powell and Dooling, 2004; Ingham 2019). We used rise-fall times of 5 ms in our 25ms long tone bursts, because initial trials showed that 1 ms rise time resulted in click artefacts at low frequencies, where the rise-fall time is less than a cycle. Because initial measurements contained onset and offset responses, we increased the rise-fall times. Despite this limitation, these data provide a reliable within-method comparison across development. Sensitivity increases monotonically with age and emerges fnrst at low frequencies, progressing toward higher frequencies. This functional trajectory matches patterns described across vertebrates (Saunders et al., 1973; Lippe, 1994; Jones et al., 2006). Hovewer it contrasts with an earlier ABR-based audiogram in zebrafnnches (Amin et al., 2007) and with the base-to-apex (high-to-low frequency) gradient of cochlear differentiation. This dissociation indicates that maturation of synaptic transmission, neural innervation, and central processing—rather than hair cell differentiation alone—limits early auditory function (Cohen and Fermin, 1978; Caus Capdevila et al., 2021). Future work using masked ABR may better resolve early frequency tuning (Brandt et al,. 2018).

ABR-derived audiograms systematically overestimate auditory thresholds relative to behavioral measures, because they require synchronous activation of large neural populations, whereas behavioral detection can rely on the most sensitive auditory nerve fnbers (Pickles, 2012). Across taxa, this offset is typically on the order of 20 dB, depending on stimulus parameters and threshold criteria (Stapells, 2000; Brittan-Powell et al., 2002). This relationship is also observed in birds, including zebra fnnches, where behavioral thresholds are consistently lower than ABR thresholds but show similar overall tuning and frequency dependence (Fig 4A). Importantly, this offset does not alter the developmental trajectory we report: even accounting for a 20-40 dB difference, early hatchlings remain far less sensitive than adults, particularly at higher frequencies. Thus, while ABR underestimates absolute sensitivity, it provides a robust and conservative measure of functional auditory maturation across development.

The postnatal maturation of hearing may also affect the onset and efficacy of vocal imitation learning. In zebra fnnches, sensory song learning starts on 25 DPH (Immelmann, 1967; Böhner,1990; Roper and Zann, 2006; Braaten, 2010; Gobes et al., 2019; Leitão and Gahr, 2024). While the auditory sensitivity to click stimuli appears be mature at 10 DPH, the response amplitude of the auditory nerve (ABR wave I) increases past 20 DPH. Furthermore, ABR features that are generated by more upstream neural populations, e.g. wave II and later, appear to go through major changes after 10 DPH (Fig 1C). These data suggest that both peripheral auditory nerve and downstream circuits of the zebra fnnch auditory system continue to develop for several weeks after hatching. In chickens, the temporal coding ability of the auditory nuclei - the ability to respond to high stimulus rates – also increases post-hatch (Saunders et al., 1973, 1974; Manley et al., 1991; Warchol and Dallos, 1990). However, we know surprisingly little about these changes in zebra fnnches in early development (Hong and Sanchez, 2018) but they seem to continue well into the sensorimotor phase of song learning. The striking co-occurrence of the onset of the sensory phase of song learning with the maturation of the auditory nerve responses suggests that the onset of song learning in zebra fnnches may be gated by the developmental maturation of the auditory system. Furthermore, nascent, internal song templates acquired prior to the full maturation of hearing function may also require updating to match the changing target.

Methods and Materials

Animals

Local, captivity-bred zebra fnnches (Taeniopygia guttata, order Passeriformes) were used in this study. Zebra fnnches were kept and bred indoors in temperature- and humidity-controlled group aviaries (100-to-200 individuals) or pairwise in breeding cages supplemented with nesting boxes. Lights in the aviaries were on a 12/12 hour light/dark photoperiod. Zebra fnnches were given water, food, cuttlefnsh bones, and nesting material ad libitum. Nesting boxes were monitored daily for accurate aging of the hatched chicks. The day when a given zebra fnnch chick was observed to have hatched in the morning was defnned to be post-hatch day 0 (0 DPH). Hatched chicks were recognized from each other by making individualized, non-invasive down feather cuts until the usage of leg bands was feasible at DPH 10. Weights for each age group are listed in Table S6. Animal experiments were approved by the Danish National Animal Experimentation Board (Dyreforsøgstilsynet) under license number 2019-15-0201-00284.

Anesthesia

Preparation

30-to-60 minutes prior to recording, 20 DPH-to-adult zebra fnnches were removed from their breeding cage or aviary, brought to the test room in a transport cage, and were fasted to minimize the reflux of fluid/food during anesthesia. Younger zebra fnnches were moved to the test room in a warmed transport box and were not fasted. Zebra fnnches were weighed prior to the induction of anesthesia.

Dosing

Anesthesia was induced with an intrapectoral injection of ketamine (Ketaminol Vet., 100 mg/ml, MSD Animal Health, NJ, USA) in combination with xylazine (Rompun vet., 20 mg/ml, Bayer Animal Health GmbH, Leverkusen, Germany). 20 DPH-to-adult zebra fnnches were administered 50 µg of ketamine and 10 µg of xylazine per gram of body weight. 2-to-10 DPH zebra fnnches were given 50 µg of ketamine and 5 µg of xylazine per gram of body weight. After one hour, a supplementary dose (75% of the induction dose) was administered. In some zebra fnnches, a clear liquid - possibly due to fluid reflux or due to increased production of oral secretions - was noticed in their mouths within 10-to-15 minutes from anesthesia induction. To avoid airways from being obstructed, a small piece of paper tissue was inserted to the side of the mouth to drain the liquid.

Auditory brainstem responses (ABRs)

Recording room

All recordings were performed by placing the zebra fnnch inside an audiometric room (T-room, C-A TEGNÉR AB, Bromma, Sweden) equipped with a self-built vibration isolation table with a padded surface. The walls of the room were padded with sound absorbing wedges (Classic Wedge 30, EQ Acoustics, UK).

Maintenance of body temperature

The body temperature of the animals was maintained by placing a heating pad warmed by circulating water (TP702, Gaymar Instruments Inc., Orchard Park, NY, USA) underneath them and placing a small cloth on top of their body. Body temperature monitoring was performed with a J-type temperature probe connected to a thermocouple measurement device (NI USB-TC01, National Instruments). The tip of the probe was coated with Sylgard (184 Silicone Elastomer Kit, Dow Europe GMBH, Wiesbaden, Germany), covered with vaseline, and was placed withing the cloaca. With 2-to-10 DPH zebra fnnches where the cloaca was too small for probe insertion the probe was placed under the wing. With this method we could keep body temperatures ±0.5°C around the target temperature during an experiment. Ambient temperature was 22°C and body temperatures for each age group are listed in Table S6. Animals were thus all 1) kept at constant temperatures (±0.5°C), 2) very close to normal body temperature (39.6-40.0°C and 3) at the same temperature (max difference: 0.6°C). All animals including very young hatchlings, thus did not suffer hypothermia that can affect ABR sensitivity (Rossi and Britt 1984; Doyle and Fria 1985). Conditions were thus consistent across all age groups, and ABR responses are thus directly comparable within this experiment.

Head placement

The head of the zebra fnnch was slightly elevated by placing it on top of wax mound (Surgident Periphery Wax, Kulzer, IN, US). The ends of a wedge-shaped notch on the wax were slightly pressed together so that the wax held the lower mandible of the zebra fnnch in place. The auricular feathers covering the left external auditory meatus in mature zebra fnnches were cut with scissors to allow unobstructed stimulation of the auditory system. Prior to electrode placement, the appropriate depth of anaesthesia was confnrmed by the lack of responsiveness to a toe-pinch.

Electrodes

Three subdermal needle electrodes were placed subcutaneously on the head of the animal in the following confnguration (Fig. 1A): (1) Active electrode: Dorsal to the left auditory meatus, pointing towards to the top of the head. (2) Inverting electrode: On top of the head along the midline, pointing towards the bill; (3) Ground electrode: Dorsal to the right auditory meatus, pointing towards to the top of the head. Of note, to compare these recordings to studies where the positions of the active and inverting electrodes are the opposite one must invert the recorded signal. The impedances of the electrodes were measured (RA4LI, Tucker-Davis Technologies, FL, USA) to be below 1 kOhm in 2-to-10 DPH zebra fnnches and below 3 kOhm in 20 DPH-to-adult zebra fnnches. Because we gathered ABRs with subdermal needles, there was no local cooling due to surgical exposure as may have been the case in sides where electrodes were places directly into the brain (e.g. Amin et al., 2007, Zevin et al., 2004).

Recording devices

Recorded evoked potentials were passed from the electrodes to a headstage (RA4LI) in series with a preamplifner (RA4PA) and with a digital signal processor (RM2, all from Tucker-Davis Technologies, FL, USA). RM2 was controlled with custom-written software QuickABR (Brandt et al., 2018; Lauridsen et al. 2021). Recorded responses were amplifned by 74 dB and digitized with a sampling rate of 25 kHz (16 bits).

Sound stimulation

Sound stimulation was generated in the RM2 controlled by QuickABR, amplifned (Cambridge Audio, Azur 740A Integrated Amplifner, London, UK), and emitted from a loudespeaker (Wharfedale Diamond 220, Wharfedale Ltd., Huntingdon, UK). The loudspeaker was string suspended 25 cm away from the animal, directly facing the left auditory meatus. Sound reflections were minimized by placing the loudspeaker at a 45° angle relative to the closest wall of the record ing room. To calibrate the loudspeaker, a ½” microphone (G.R.A.S. microphone type 26AK, G.R.A.S., Holte, Denmark) was suspended at the position of the head of the animal and connected to a microphone amplifner (G.R.A.S. Power Module Type 12AA, Gain 0, Linear fnltering, Direct mode select = Ampl.). The microphone itself was calibrated using an acoustical calibrator (type 4321, Output: 1kHz at 94 dB re. 20 µPa, Brüel & Kjær). Calibration of the loudspeaker was controlled by QuickABR.

Stimuli

ABRs were elicited with broadband clicks (20 µs) and 25 ms-long tonebursts at 0.25 kHz, 0.5 kHz, 1 kHz, 2 kHz, 3 kHz, 4 kHz, 6 kHz, and 8 kHz. To minimize spectral splatter caused by abrupt on/offsets of stimuli, the rise/fall time (5 ms, cos2-gated) of the tonebursts was selected to accommodate at least one full cycle of the stimulus with the lowest frequency. To further increase the frequency specifncity of the stimulus, the plateau time (15 ms) was selected to allow at least 3 full cycles of the stimulus with the lowest frequency of 250 Hz. Using a constant duration across frequencies ensures comparable stimulus energy across frequencies.

Additionally, 5 ms-long tone pips (1 ms rise/fall time, cos2-gated) were collected from adult zebra fnnches. Stimuli were sampled at 25 kHz and at a repletion rate of 25 Hz (clicks) and 3Hz (pips and bursts) to collect average responses consisting of 400 individual responses. ABR signal-to-noise ratio improves with the square root of the number of averages. Increasing from 400 to 1000 sweeps would reduce thresholds by at most ∼4 dB. Every second presented individual stimulus was inverted to reduce microphonic effects. The maximal sound pressure level (SPL) used in this study was selected to be 95 dB SPL as it can be viewed as being physiologically relevant. Click amplitudes are stated as the root mean square (RMS) dB value of a sinusoid with the same peak amplitude as the click (peak equivalent dB).

Recording paradigm

The stability of the recorded responses was monitored throughout the recording session by recording a 95 dB SPL click-evoked ABR periodically between each stimulus type. A specifnc stimulus recording was accepted into further analysis if the amplitude of the fnrst click-evoked ABR wave had changed less than 25% from the most temporally stable amplitude of that recording session. To minimize test order bias, the recording order of stimuli with varying frequencies was randomized. Three animals were killed with an overdose of ketamine at the end of the recording session to verify the lack of response potentials post mortem.

ABR recovery from paired clicks

If consecutive clicks are played back at a too high rate, this can suppress the ABR response (Brittan-Powell and Dooling, 2004). To test the effects of stimulus repetition rate on the ABR, fnve adult zebra fnnches were presented at 6 Hz with paired-click and single click stimuli (Henry et al., 2011). The time between the fnrst and second click (inter-click interval) was varied at 1 ms, 2 ms, 4 ms, 8 ms, 16 ms, 32 ms, and 48 ms. Each ABR was averaged from a total of 512 stimulus representations. The derived ABR to the second click was obtained by point-by-point subtraction of the single click stimulus ABR from the paired click stimulus ABR. To calculate the percentage of ABR recovery, the amplitude of the ABR wave I of the derived, second response was divided by the amplitude of the ABR wave I to the single click stimulus. These data show that the click rate of 25 Hz we used did not affect the magnitude of the ABR response (Figure S3). The burst rate of 3 Hz is a low presentation rate and is not expected to affect the ABR amplitude.

Egg vibration measurements with laser doppler vibrometry

Zebra fnnch eggs (N = 5) were removed from their nest and placed in an upright position on a dental wax covered platform at the centre of an anechoic room. The anechoic room was custom-made with its walls and floor covered with 30 cm rockwool wedges. The sample platform was mounted on a heavy tripod which also held a laser Doppler vibrometer (OFV-5000 with OFV-505 sensor, Polytec, Waldbronn, Germany) attached on a manipulator. The laser was pointed at the centre of the side of the egg. Strong reflections were directly obtained from the egg surface so adding reflectors was not necessary. Sound-induced vibrations of the egg were induced with a speaker (JBL IG, Northridge, CA, USA) placed next to the laser vibrometer. A microphone (½ inch G.R.A.S., Brüel & Kjær) was placed few millimeters above the egg and was used for speaker calibration. The egg was stimulated with sound frequency sweeps (0.25-to-10 kHz) at 84 dB SPL. The sweeps were generated with Tucker-Davis system 2 hardware (Tucker-Davis Technologies, TDT, Alachua, FL, USA) controlled by custom-written software DragonQuest (Christensen-Dalsgaard and Manley, 2008).

The signal was deconvoluted by dividing the spectrum of the sweep with the measured transfer function of the speaker. The analog laser signal of the laser vibrometer was digitized using an A–D converter (AD2, TDT) and averaged over 100 presentations. During recordings, the laser signal was monitored using an oscilloscope and a vibrometer controller (Polytec OFV-5000). Stimulation and recording were controlled by DragonQuest.

Data analysis and visualization

Data was visualized and analysed with custom scripts in MATLAB R2022B (Mathworks). Figures were edited with Illustrator CS6 (Adobe).

Threshold determination

For ABR threshold detection, the stimulus level was lowered from the maximal value in 5 dB steps. For the reported thresholds to reflect the average sensitivity of zebra fnnches capable of detecting the given stimulus, only zebra fnnches that displayed a detectable ABR (both visually and above the signal-to-noise ratio of 2) at the maximal stimulation level are included in the analysis. The response threshold was designated to be 2.5 dB below the last observable response-generating stimulus level (arrow in Fig 1F).

Signal-to-noise ratio calculation

The recording prior to stimulus onset was used to calculate the noise RMS. Signal-to-noise ratio (S/N) was determined by dividing the RMS value of the signal (blue window) by the RMS of the recording prior to stimulus onset (red window, Figs. 1D; 2A,B and Fig. S2A). We used an automated criteria detection of thresholds to remove potential experimenter bias and checked the automatic thresholds by visual inspection.

Response latency and amplitude calculation

Response latency was calculated as the time difference between the fnrst negative deflection of the ABR (designated as wave I) and the arrival time of the sound at the ear (Fig 1C). The amplitude of the wave I (reflecting the activity of the auditory nerve) was calculated as the potential difference between the fnrst negative peak and the following positive trough that together formed the most prominent component of the response (Fig 1C). Of note, a second negative peak/deflection between the fnrst peak and the trough existed. This peak appeared to merge with the fnrst negative peak at soft SPLs as reported for ABRs of bald eagles and red-tailed hawks (McGee et al., 2019).

Statistics

To test for main effects, we used linear mixed effects models. We sequentially compared models by adding hypothesised effects as fnxed effects one-by-one, starting from a null model and ending with full interaction models. For all age comparisons, age was treated as an ordinal rather than a continuous variable to account for discrete timepoints of measurements and to allow for post-hoc pairwise comparisons between ages. All models had the individual animals as random effect. We used chi-square tests, Akaike Information Criterion (AIC), and Bayes Information Criterion (BIC) as criteria for model comparison. For models with signifncant main effects, we then used estimated marginal means for post-hoc comparisons. We used Bonferroni p-value correction to account for multiple comparisons. In all fngures, we denote signifncance levels as follows: *, p(Bonf.) = 0.05; **, p(Bonf.) = 0.01; ***, p(Bonf.) = 0.001. All statistical testing was carried out in R 4.2.1 (R Core Team, 2024) We used the lme4 package (Bates et al., 2015) for mixed effects modelling and emmeans package (Lenth, 2024) for post-hoc comparisons.